---
title: "The real option value of multistage agricultural research for development"

author:
  - name: Benjamin Schiek
    email: b.schiek@cgiar.org
    affiliation: CIAT
    correspondingauthor: true

address:
  - code: CIAT
    address: Policy, Inclusion, and Socioeconomic Analysis unit (PISA), Alliance of Bioversity International and CIAT, Km 17 Recta Cali-Palmira, Palmira, Colombia 763537

#abstract: |
  # Agricultural research for development (AR4D) donors face austere fiscal outlooks that restrict their ability to commit to high impact, high risk, agricultural research projects with long time horizons. This results in a series of short-term, piecemeal funding arrangements that inhibit the AR4D community's ability to conduct the disruptive research necessary to meet the Sustainable Development Goals. Real option valuation (ROV) presents a mechanism offering donors the flexibility they require to justify long term research funding commitments. However, adaptation of existing ROV methods to the AR4D context is complicated by the financial and corporate underpinnings of these methods, which are inconsistent with the non-market character of most AR4D projects. Adaptation is also complicated by the multistage structure of most AR4D projects. Here a multistage ROV method is derived from theoretical underpinnings that are consistent with the AR4D context. The method is applied to evaluate a real, four stage potato research project.

output:
  rticles::plos_article:
    keep_tex: true

bibliography: AR4Drealoptions.bib
csl: https://www.zotero.org/styles/plos

header-includes:
  #- \renewcommand{\theequation}{(\arabic{equation})}
  - \microtypesetup{expansion=false}
  - \usepackage{lmodern}
  #- \numberwithin{equation}{section}
  - \usepackage{float}
  - \floatplacement{figure}{h}
  - \usepackage{caption}
  - \usepackage{setspace}
---

<!-- \pagebreak -->

<!-- Funding acknowledgement: -->

<!-- This work was carried out under the CGIAR Science Programs on Policy Innovations and Breeding For Tomorrow. We would like to thank all funders who supported this research through their contributions to the CGIAR Trust Fund (www.cgiar.org/funders). -->

\pagebreak

# Abstract

Agricultural research for development (AR4D) donors face increasingly austere fiscal outlooks that restrict their ability to commit to high impact, high risk, agricultural research projects with long time horizons. This results in a series of short-term, piecemeal funding arrangements that inhibit the AR4D community's ability to conduct the disruptive research necessary to meet the Sustainable Development Goals. Real option valuation (ROV) presents a potential mechanism offering donors the flexibility they require to justify long term research funding commitments. However, adaptation of existing ROV methods to the AR4D context is complicated by the financial and corporate underpinnings of these methods, which are inconsistent with the non-market character of most AR4D projects. Adaptation is also complicated by the multistage structure of most AR4D projects, since existing ROV methods are mostly limited to single stage projects. Here I present a multistage ROV method derived from novel theoretical underpinnings that are consistent with the AR4D context. As an illustrative example, I apply the derived model to evaluate a real, four stage potato research project.

<!-- Keywords: -->

  <!-- - Real option value -->
  <!-- - Multistage research for development -->
  <!-- - Far-from-market innovation -->
  <!-- - Risk neutral valuation -->

\pagebreak

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      #fig.pos = 'H',
                      dev = c("png", "tiff"),
                      dpi = 300)
# null-coalescing helper
`%||%` <- function(a, b) if (!is.null(a)) a else b

# submission-mode figure hook
knitr::knit_hooks$set(plot = function(x, options) {
  if (is.null(options$fig.cap)) return(NULL)

  # choose a label (use chunk name by default)
  lab <- options$label %||% ""
  if (is.null(lab) || lab == "") lab <- options$label

  paste0(
    # advance counter and label
    "\\refstepcounter{figure}\n",
    if (nzchar(lab)) paste0("\\label{fig:", lab, "}\n") else "",
    # use LaTeX's \thefigure for both occurrences so numbers stay in sync
    "\\begin{center}\\textit{[Fig \\thefigure\\ about here]}\\end{center}\n\n",
    # left-aligned caption: bold only "Fig n."
    "\\noindent\\textbf{Fig \\thefigure.} ", options$fig.cap, "\n\n\\par\\vspace{1em}\n\n"
  )
})

library(tidyverse)
library(flextable)
library(patchwork)
library(kableExtra)

label_size <- 2.5
smallLabel_size <- 2
title_size <- 8
subtitle_size <- 7
legendText_size <- 6
axisText_size <- 6
axisTitle_size <- 7
facetTitle_size <- 7

```

\onehalfspacing

# Introduction

Agricultural research for development (AR4D) projects tend to be high reward, but also high risk, propositions with long time horizons. Donors are attracted by the high potential payoff of such projects, but face increasingly austere fiscal outlooks that restrict their ability to engage in long term, risky commitments. For researchers, the resulting uncertainty in the long term funding outlook incentivizes work in the direction of low risk, modest objectives that are achievable in the near term. Such research has its place; but the Sustainable Development Goals (SDGs) call for higher impact, riskier, longer term research as well. Real options valuation (ROV), a method of project appraisal adapted from financial analysis, presents a potential enabling mechanism by which to secure long term funding for disruptive research while also guaranteeing donors the fiscal flexibility they require to justify such commitments.

The notion of a real option, first introduced by Myers [-@myers1977determinants], is born of the analogous concept of a type of financial option known as the European call option. The holder of a European call option on an underlying security (a stock, for example) has the right, but not the obligation, to purchase the security for a predetermined price---called the "strike" or "exercise" price---at the end of a certain time period. Analogously, donors who "buy" a real option on an underlying project commit to funding the project for a certain time period, at the end of which they have the right, but not the obligation, to continue funding the subsequent stage of the project. In the case of research projects, for example, the subsequent stage at the end of time period $T_0$ might be the release and outscaling of a new crop variety developed during the main research stage. The value of such a real option ($ROV_1$), as evaluated at the time of the investment decision ($t = 0$), may be expressed mathematically as follows.

\begin{equation}
ROV_1(t)|_{t = 0} = e^{-r T_0} E[\max(x(T_0) - K_0, 0)]|_{t = 0}
\label{eq:basic}
\end{equation}

Where $x(T_0)$ is project net present value (NPV) at the end of the research period $T_0$, $K_0$ is the cost of funding the subsequent project stage (e.g., release and outscaling of the finished research product), and $r$ is the discount rate. By contrast, the comparable conventional appraisal of the project is calculated as

\begin{equation}
e^{-r T_0} (E[x(T_0)]|_{t = 0} - K_0)
\label{eq:conv}
\end{equation}

To characterize this as the conventional "NPV approach", as is sometimes done in the ROV literature [e.g., @koppl2013real], can be confusing since $x(t)$ is already defined as project NPV at any time $t$ (net of $K_0$), and appears in the calculation of ROV. To be clear, then, in the present article, the conventional manner of project appraisal in Eq \eqref{eq:conv} is referred to as the "Cost Benefit Analysis (CBA) approach" or "CBA basis"; whereas "project NPV", denoted $x(t)$, refers to the expected value at time $t$ of the net benefit stream resulting from release and outscaling of the research product after time $T_0$, net of any costs incurred at or prior to time $T_0$. As in other treatments of ROV in research contexts [@dixit1994investment], the present work does not address the complexities involved in the evaluation of $x(t)$ itself; which include, e.g., non-market valuation of public goods, payment for environmental services, public health outcomes, and the valuation of other intangibles (see Freeman et al. [-@freeman2014measurement] for an introduction to the topic). For present purposes, it is assumed that an evaluation of $x(t)$ exists, typically as the result of an ex-ante impact assessment.

On an ROV basis, the investment $K_1$ required to implement the research is justified if

\begin{equation}
e^{-r T_0} E[\max(x(T_0) - K_0, 0)]|_{t = 0} > K_1
\end{equation}

On the CBA basis, the criterion is

\begin{equation}
e^{-rT_0} (E[x(T_0)]|_{t = 0} - K_0) > K_1 \:\: \rightarrow \:\: \text{invest in project}
\end{equation}

ROV is typically higher than the value $e^{-rT_0} (E[x(T_0)]|_{t = 0} - K_0)$, such that a project that would normally be rejected on the conventional CBA basis could be accepted on an ROV basis. This is especially true in the case of high risk, high reward projects. A comprehensive introduction to ROV and the main developments since its inception is given by Trigeorgis [-@trigeorgis1999real] and Trigeorgis and Tsekrekos [-@trigeorgis2018real]. For an introduction to ROV in R&D contexts, see Doctor et al. [-@doctor2001managing], and Newton et al. [-@newton2004real]. The agricultural ROV literature focuses on diverse aspects of farm or firm level investment options and technology adoption [@koppl2013real; @spiegel2020stochastic; @perez2024valuation; @smith2018real; @biancardi2024valuation]; but has not yet considered the ROV of multistage research from the perspective of donors and institutions.

Real options contracts potentially bridge the divide between donor risk aversion and the risk tolerance required to achieve high impact AR4D because they are mutually beneficial for both donors and researchers alike. From the donor's perspective, a real option contract is a way to tentatively lay claim to promising, but distant and uncertain, research results, while at the same time guaranteeing easy extraction from underperforming projects. The resulting fiscal agility then translates into longer term funding commitments, which is music to the ears of the research community.

It is worth noting, moreover, that many donors already acquire an unacknowledged option value in their conventional research funding arrangements. That is to say, nowadays, most AR4D contracts are signed for no more than 1-5 years, at the end of which the contract comes up for renewal or cancellation, pending progress up to that point. By reserving the right to discontinue funding at specific time intervals, the donor effectively acquires option value. The introduction of ROV would thus merely formalize an already existing practice, in such cases, and more accurately appraise the real value acquired by the donor in exchange for their investment. By the same token, the default CBA methods currently in place neglect to price in the donor's option value, and thus understate the real value acquired by the donor.

ROV methods have so far developed primarily in the context of corporate projects. A simple extension of this work to the AR4D context is complicated, if not made impossible, by the fundamental difference between the two contexts: corporate projects unfold in market or near-market settings, where much of the original option value theory developed in finance may be invoked. AR4D projects, on the other hand, typically unfold in the midst of market failures, where the underlying assumptions of such theory do not apply. In particular,

1) AR4D project NPV is not well approximated by a linear combination of price movements (the "replicating portfolios" assumption)

2) AR4D project risk cannot be hedged away by trading securities (the "complete markets" assumption)

3) AR4D donors are not willing to take on any level of risk for a given expected return ("risk neutral valuation").

Extension of existing ROV methods to the AR4D context is also complicated by the multistage nature of most AR4D projects. For example, many AR4D plant breeding projects involve a preliminary greenhouse and/or minimal field trial stage, followed by larger multi-location, multi-season field trials [@covarrubias2022breeding]. In transgenic projects, additional toxicological testing is required, followed by the compilation of a regulatory dossier and application for deregulation before the national competent authority in the target countries where the new technology is to be released [@schiek2016demys]. Finally, there is the release, distribution, and outscaling stage, which involves careful orchestration of complex networks of stakeholders and public-private partnerships.

Donors invest in each of these stages one at a time, consecutively, with investment in the subsequent stage contingent upon successful completion of the current stage. The donor's investment in stage 1 of a 3-stage project, then, is analogous to purchasing an option on stage 2---i.e., the right but not the obligation to invest in stage 2---which is in turn an option on stage 3. The ROV of stage 1 of such a project is the value of the option to invest in stage 2. This may be expressed as follows.

\begin{equation}
ROV_3(t)|_{t = 0} = e^{-r T_2} E[\max(ROV_2(T_2) - K_2, 0)]|_{t = 0}
\label{eq:eg3stage}
\end{equation}

Where $ROV_2(T_2)$ is the ROV of stage 2 evaluated at $T_2$ (the end of stage 1), $K_2$ is the cost of investing in stage 2 upon successful completion of stage 1; and stage 1 is considered successfully completed if $ROV_2(T_2) > K_2$. And $ROV_2(T_2)$, in turn, is defined

\begin{equation}
ROV_2(T_2) = e^{-r T_1} E[\max(ROV_1(T_1) - K_1, 0)]|_{t = T_2}
\end{equation}

Where $ROV_1(T_1)$ is the ROV of the third and last stage of research evaluated at $T_1$ (the end of stage 2), $K_1$ is the cost of investing in stage 3 upon successful completion of stage 2; and stage 2 is considered successfully completed if $ROV_1(T_1) > K_1$. And $ROV_1(T_1)$, in turn, is the ROV of the last stage---the value of the option to invest in release/outscaling/marketing of the finished research product with NPV $x(T_0)$---as defined in Eq \eqref{eq:basic}.

If $ROV_3(0)$ is greater than the investment $K_3$ required to implement stage 1, then investment in stage 1 is justified. The comparable CBA criterion for any $n$-stage research project would be

\begin{equation}
e^{-rT_0} E[x(T_0)]|_{t = 0} - \sum_{i = 0}^{n - 1} e^{-r T_i} K_i > K_n \:\: \rightarrow \:\: \text{invest in project}
\label{eq:cbaCrit}
\end{equation}

Where $n = 3$ in this particular example.

Existing ROV methods, meanwhile, focus primarily on single stage options. A niche of two-stage studies [e.g., @biancardi2024valuation] adapt the "compound option" model pioneered by Geske [-@geske1979valuation] in financial markets; and Cassimon et al. [-@cassimon2004valuation] extend this framework to multistage research investments (i.e., investments of arbitrary stage number) for the valuation of pharmaceutical trials. Again, such adaptations are not readily extensible to the AR4D context due to their deep roots in financial theory. A separate strand of the literature is multistage in the sense that it assesses the optimal timing of investments across multiple consecutive decision points [e.g., @mombello2023valuing; @hassi2022compound]. This is an adaptation of a different kind of financial call option known as the American call option, which may be exercised at any moment up to a given time limit [@hull9thEdition]. Such methods are inapplicable for present purposes due both to their financial theoretical underpinnings and because time horizons between investment decisions in the AR4D setting are generally fixed. Breaking from the financial lineage entirely, Belz et al. [-@belz2025real] introduce a novel mixed-integer programming approach that accommodates multiobjective, non-financial benefit streams for the assessment of portfolios of public and private R&D projects. While this presents an interesting alternative for portfolio level assessment, the present work is concerned with ROV at the level of individual projects. Furthermore, it is concerned with research investments of arbitrary stage number, whereas Belz et al.'s discrete formulation incurs major computational challenges beyond the two-stage case.

To redress this lacuna, here I derive a method for evaluating the ROV of multistage projects in a non-market setting (e.g., Eq \eqref{eq:eg3stage}), with particular reference to AR4D as the motivating applied context. Rather than build upon the existing ROV groundwork cited above, which is largely inherited from the financial and corporate contexts, I demonstrate why that groundwork is incompatible with the AR4D context, and replace it with suitable foundations. More specifically, in place of strained arguments that would construe project NPV as a traded good in an efficient market, I show how a tractable multistage ROV model for AR4D research products follows from the definition of NPV itself, and from the assumption that the time evolution of project NPV follows a geometric Brownian motion (gBm). Because the gBm model is widely eschewed in the ROV literature, the main derivation of the proposed multistage model is preceded by a substantial defense of this methodological choice. To illustrate the proposed method's utility, the main derivation is followed by a hypothetical use case to evaluate the ROV of stage 1 research of a real 4-stage AR4D project (Late Blight resistant potato).

It is important to distinguish the work presented here from a separate "real options approach" that examines the optimal delay, from a social planner’s perspective, of releasing a completed transgenic agricultural R&D product under environmental uncertainty [@scatasta2006irreversibility; @demont2004biodiversity; @wesseler2007maximum; @demont2005irreversible]. That line of work builds on the American-style option-to-delay framework developed by Dixit and Pindyck [-@dixit1994investment] in conjunction with the concept of "quasi-option value" developed in environmental economics [@fisher2000investment]. By contrast, the present work concerns the valuation of sequential, stage-gated investments in research that has yet to be undertaken, modeled as European-style options with fixed-horizon investment cycles, as formalized in Eq \eqref{eq:basic}.

# The incompatible financial underpinnings of existing ROV approaches

## AR4D products are not traded goods in efficient markets

The ROV literature cited above has developed in the context of corporate projects aiming to develop new goods or services to be marketed for a profit. In such a market or near-market setting, a financial argument known as "replicating portfolios" may be invoked [@brennan1985evaluating; @koller2010valuation; @amram1998real]. Methodologically, the replicating portfolios argument means that corporate project NPV is well approximated ("replicated") by a linear combination of relevant prices. This greatly facilitates the estimation of project NPV and risk---both of which are required for ROV---as they may be calculated at any time from publicly available price information.

AR4D projects, on the other hand, are born of market failures, where no such argument may be invoked; and so AR4D project NPV and risk must be calculated the "hard way", by ex-ante impact assessment. Some might contend that the recent emergence of carbon and green bond markets [@tolliver2019green; @bhutta2022green] paves the way for a replicating portfolios analogue in the AR4D context. However, in addition to green outcomes, AR4D projects typically focus on health, nutrition, and socioeconomic outcomes for marginalized populations, the respective values of which, almost by definition, are not reflected in any market. And, even if they were---even if there were an "SDG bond" mapped to every AR4D impact area---this might be useful in the evaluation of project NPV, but of very little use in the assessment of project risk, since the main source of AR4D risk is not "price risk" (tragically, there is little risk of the bottom falling out of the SDG "market" in our lifetimes), but rather technical risk, i.e. the risk that the proposed technology will not function or scale as expected, due either to technical difficulties encountered in the research itself, or to unforeseen adversity in the enabling environment (e.g., government policy shifts, deregulation issues, etc.).

Even in the corporate ROV context, the replicating portfolios argument is met with considerable skepticism [@borison2005real]; and proponents of the argument acknowledge that its validity is limited to a narrow range of project types, which generally does not include R&D projects [@schwartz2013real]. There is strong methodological incentive to invoke the argument because the fundamental equation governing all types of option (and all financial derivative) valuation---i.e., the Black-Scholes partial differential equation (PDE)---is formally derived from an assumption in financial markets known as the "no-arbitrage rule", which requires that the underlying project NPV be interpretable as a traded good in an efficient market [@hull9thEdition].

To preserve the Black-Scholes PDE in cases where the replicating portfolios argument may not be invoked, some corporate ROV practitioners invoke the "complete markets" assumption as a second-best alternative [@majd1987time; @pennings1997option; @schwartz2013real]. Under the complete markets assumption, project NPV cannot be modeled as a linear function of prices, but the Black-Scholes PDE is nonetheless preserved because "all risks can be hedged by trading securities" [@pennings1997option]. As just mentioned, the main source of AR4D project risk is technical risk associated with non-market factors. Clearly, then, AR4D risk cannot be hedged by trading securities; and so there are no grounds in the AR4D context on which to invoke complete markets.
<!-- i.e. where project NPV may not be replicable as a traded good, -->

More problematically still, the formal derivation of the Black-Scholes PDE from either the replicating portfolios or complete markets assumptions leads to a celebrated corollary known as "risk-neutral valuation", which implies that "investors do not increase the expected return they require from an investment in order to compensate for increased risk" [@hull9thEdition]. Clearly, AR4D donors are not risk neutral. They are willing to bet on a high risk project only if there is a sufficiently high expected return on the investment.

In the derivation of the AR4D ROV formula presented farther below, these problematic foundations are replaced with the comparatively simple assumption that future values of project NPV are lognormally distributed. The derivation also requires that the expected growth rate of project NPV be equal to the discount rate, but this follows trivially from the definition of project NPV itself, and thus does not have to be assumed.

For present purposes, the assumption of lognormally distributed project NPV is tantamount to assuming that project NPV evolves as a gBm, a notion that is eschewed in the financial and corporate contexts in favor of more complex stochastic models [@merton1976option, @tankov2003financial]. Before proceeding to the derivation of the multistage AR4D ROV formula, then, it is first necessary to introduce, contextualize, and defend the choice of gBm as a suitable model of the time evolution of project NPV.

## In defense of gBm as a reasonable model of the time evolution of AR4D project NPV \label{sec:gBmDefens}

In settings such as AR4D where time series of asset values generally do not exist, a suitable choice of stochastic model of the time evolution of project NPV may nonetheless be guided deductively by consideration of the size distribution of future values of NPV, or the size distribution of changes in NPV per relevant small time increment (e.g., an annual quarter in the context of 10-20 year AR4D projects). Changes in project NPV over such intervals are driven by changes in underlying component quantities---the probability that a research target is achieved, the likely rate of technology adoption, the estimated magnitude of welfare impacts on target populations, and so forth. These changes operate multiplicatively on project value. A change in adoption potential, for instance, scales the entire expected benefit stream upward or downward by some factor rather than adding or subtracting a fixed amount. This multiplicative structure implies that proportional changes in NPV---equivalently, changes in log NPV---are the natural unit of analysis, rather than absolute changes in NPV levels. When such proportional changes are individually small, collectively numerous, and approximately independent, log NPV changes over small time increments may plausibly be approximated as normally distributed. Normally distributed log NPV changes, in turn, are the distributional signature of a gBm process. The left panel of Fig \ref{fig:NPVevol} depicts a simulated gBm trajectory generated through the accumulation of such proportional changes.
<!-- A natural starting point in this deductive process is....A natural starting point is to ask what kind of changes in AR4D project NPV are most plausible on basic logical grounds. -->
<!-- Many in the AR4D audience may be unfamiliar with stochastic models, but well versed in statistics. For purposes of exposition, the choice of stochastic model of project NPV time evolution may be roughly thought of as the choice of probability density function of future values of project NPV, or of the size distribution of changes in project NPV per small time increment. Lognormally distributed percentage changes in project NPV per annual quarter imply a gBm stochastic process. An example of gBm is depicted in the left panel of Fig \ref{fig:NPVevol}. -->
<!-- Many in the AR4D audience may be unfamiliar with stochastic models, but well versed in statistics. For purposes of exposition, the choice of a stochastic model of project NPV time evolution may be thought of as the choice of the distribution of changes in project NPV over small time increments, or equivalently, of the implied distribution of future values. Because long, high-frequency time series on AR4D project value are generally unavailable, this choice must be guided primarily by deductive reasoning. -->

```{r, include=FALSE}
gbmFun <- function(tau = 40, m = 0.001, s = 0.003, x0 = 1,
                   randVec = NULL, seed = NULL) {
  if(is.null(randVec)){
    if(!is.null(seed)){set.seed(seed)}
    randVec <- rnorm(tau)
  }
  epsilon <- randVec
  lx <- c(); lx[1] <- log(x0)
  drift <- (m - s * s / 2)
  for(t in 2:tau){
    dBt <-  s * epsilon[t]
    lx[t] <- lx[t - 1] + drift + dBt
  }
  x <- exp(lx)
  return(x)
}
# m <- 0.001
# s <- 0.003
# tau <- 400
# x0 <- 1
# x <- gbmFun(tau, m, s, x0, randVec = NULL)
# df_plot <- data.frame(t = 1:tau, x)
# gg <- ggplot(df_plot, aes(x = t, y = x))
# gg <- gg + geom_line()
# acf(diff(log(x)))
# hist(diff(log(x)))
# shapiro.test(diff(log(x)))
#---------------------------------------------------------------------------
# Algorithm 6.2 in Tankov 2003 Financial modeling with jump processes
compoundPoisFun <- function(tau, lambda, m_y = 0, s_y = 1,
                            seed = NULL, maxiter = 500){
  if(!is.null(seed)){set.seed(seed)}
  N <- rpois(1, lambda * tau)
  U <- runif(N) * tau
  U <- round(U)
  n_iter <- 0; flag <- 1
  while(flag == 1){
    n_iter <- n_iter + 1
    U <- runif(N) * tau
    U <- round(U)
    if(sum(duplicated(U)) > 0 & n_iter <= maxiter){
      flag <- 1
    }else{
      flag <- 0
      if(n_iter == maxiter){
        print("Reached maxiter without generating duplicate-free event times vec. Dropping duplicates.")
        U <- U[-which(duplicated(U))]
      }
    }
  }
  Tt <- U[order(U)]
  J <- exp(rnorm(N, m_y, s_y)) - 1
  #---
  cumJ <- cumsum(J)
  # For explicit modeling of Yt per time step
  Yt <- rep(0, tau)
  Yt[Tt] <- cumJ
  #---
  Tt <- c(0, Tt)
  J <- c(0, J)
  if(Tt[length(Tt)] != tau){
    Tt <- c(Tt, tau)
    J <- c(J, 0)
  }
  cumJ <- cumsum(J)
  df_step <- data.frame(Tt, cumJ)
  #--------------------------
  nEvents <- length(Tt)
  cumYt <- c()
  for(i in 1:(nEvents - 1)){
    tStart <- Tt[i]
    tFin <- Tt[i + 1] - 1
    cumYt[tStart:tFin] <- cumJ[i]
  }
  cumYt[tau] <- cumYt[tFin]
  df_Yt <- data.frame(t = 1:tau, Yt, cumYt)
  
  list_out <- list(df_step, df_Yt)
  return(list_out)
  
}
#----------------------------------------------------------------------------
FractDim<-function(Data,graphon=FALSE) {
  X=Data;N=length(X);
  jstart=10;jend=floor(10*(log10(N)-1));
  kvec=c(1:4,floor(2^(c(jstart:jend)/4)));
  indkend=length(kvec);
  k=c()
  AvgLmk=c()
  err=c()
  for(indk in 1:indkend)
  {
    k=kvec[indk]
    Xend=c()
    Xsum=c()
    Lmk=c()
    for(m in 1:k)
    {
      Xend=floor((N-m)/k)
      Xsum=sum(abs(X[m+c(1:Xend)*k]-c(0, X[m+c(1:(Xend-1))*k])))
      Lmk[m]=1/k*1/k*(N-1)/Xend*Xsum
    }
    AvgLmk[indk]=mean(Lmk)
    #  err[indk]=sd(log(Lmk))
  }
  x<-log(kvec)
  y<-log(AvgLmk)
  q<-lm(y~x)
  slope<-q$coefficients[2]
  yintcept<-q$coefficients[1]
  yfit<-x*slope+yintcept
  FrDim <- -slope
  avgRes <- mean(abs(q$residuals))
  if(graphon==TRUE)
  {
    plot(x,y,main="If linear then fractal, w/Fr. Dim = (-)slope",xlab="Ln(k)",ylab="Ln(length of curve with interval k)")
    z<-line(x,yfit);abline(coef(z),col='blue');z<-NULL
    #z<-line(x,y);abline(coef(z),col='blue');z<-NULL
  }
  #z<-line(x,y);qq=coef(z)
  #yintcept=qq[1]
  #FrDim=-qq[2]
  return(c(FrDim, avgRes, yintcept))
}
#===========================================================================
m <- 0.005
s <- m * seq(5.5, 50, length.out = 3)
tau <- 48
x0 <- 1
#rn <- round(runif(1) * 1000)
#rn <- 905
#rn <- 517
# rn <- 340
# set.seed(rn)
# tau <- 12 * 4
lambda <- 0.1


#----------------------------------------------------------------------------
# Generate and plot geometric Brownian movement example
#gBm_seed <- 10^4 * round(runif(1), 4)
#gBm_seed <- 2482
#gBm_seed <- 7781
gBm_seed <- 1029
#randVec <- coloredNoise(N = tau, alpha = 1, scaleIt = T)
#acf(randVec)
list_df <- list()
list_spec <- list()
facet_labels <- c()
for(i in 1:length(s)){
  x <- gbmFun(tau, m, s[i], x0, randVec = NULL, seed = gBm_seed)
  df_x <- data.frame(t = 1:tau, x, s = as.character(s[i]))
  list_df[[i]] <- df_x
  o <- spectrum(x, plot = F)
  df_gbmSpec <- data.frame(lfreq = log(o$freq), lpwr = log(o$spec))
  df_gbmSpec <- df_gbmSpec[-1, ]
  df_gbmSpec$s <- as.character(s[i])
  list_spec[[i]] <- df_gbmSpec
  
  mod <- lm(lpwr ~ lfreq, df_gbmSpec)
  # summary(mod)
  alpha <- round(as.numeric(coefficients(mod)[2]), 2)
  # yint <- as.numeric(coefficients(mod)[1])
  # df_out <- as.data.frame(broom::glance(mod))
  # adjR2 <- round(df_out$adj.r.squared, 2)
  # N <- df.residual(mod)
  this_facet_label <- paste0("Slope = ", alpha, " fd = ", round(FractDim(x)[1], 2))
  facet_labels[i] <- this_facet_label
}
#---
df_plot <- as.data.frame(do.call(rbind, list_df))
colnames(df_plot)[1:2] <- c("Time", "Project NPV")
df_plot <- subset(df_plot, s == s[2])
gg <- ggplot(df_plot, aes(x = Time, y = `Project NPV`))
gg <- gg + geom_line()
#gg <- gg + facet_wrap(~s, scales = "free_y", ncol = 1)
gg <- gg + theme_bw()
gg <- gg + theme(axis.text = element_blank(),
                 axis.title = element_text(size = axisTitle_size))
gg_gbm <- gg
dfPlot_gBm <- df_plot
#---
df_plot <- as.data.frame(do.call(rbind, list_spec))
colnames(df_plot)[1:2] <- c("Logged frequency", "Logged power spectral density")
df_plot <- subset(df_plot, s == s[2])
#names(facet_labels) <- s
gg <- ggplot(df_plot, aes(x = `Logged frequency`, y = `Logged power spectral density`))
gg <- gg + geom_smooth(method = lm, se = F)
gg <- gg + geom_line()
# gg <- gg + facet_wrap(~s, ncol = 1,
#                       labeller = labeller(s = facet_labels))
gg <- gg + theme_bw()
gg <- gg + theme(axis.text = element_blank(),
                 axis.title = element_text(size = axisTitle_size))
gg_gbmSpec <- gg
#----------------------------------------------------------------------------
# Generate and plot compound Poisson process example
#cPp_seed <- 10^4 * round(runif(1), 4)
#cPp_seed <- 6819
#cPp_seed <- 4880
#cPp_seed <- 7656
cPp_seed <- 2558
list_df <- list()
list_spec <- list()
facet_labels <- c()
for(i in 1:length(s)){
  #s_y = 0.3
  list_out <- compoundPoisFun(tau, lambda, m_y = m,
                              s_y = s[i], seed = cPp_seed,
                              maxiter = 500)
  df_x <- list_out[[1]]
  df_x$s <- as.character(s[i])
  list_df[[i]] <- df_x
  
  df_Yt <- list_out[[2]]
  o <- spectrum(df_Yt$cumYt, plot = F)
  df_spec <- data.frame(lfreq = log(o$freq), lpwr = log(o$spec))
  df_spec <- df_spec[-1, ]
  df_spec$s <- as.character(s[i])
  list_spec[[i]] <- df_spec
  
  mod <- lm(lpwr ~ lfreq, df_spec)
  # summary(mod)
  alpha <- round(as.numeric(coefficients(mod)[2]), 2)
  # yint <- as.numeric(coefficients(mod)[1])
  # df_out <- as.data.frame(broom::glance(mod))
  # adjR2 <- round(df_out$adj.r.squared, 2)
  # N <- df.residual(mod)
  this_facet_label <- paste0("Slope = ", alpha, " fd = ", round(FractDim(df_Yt$cumYt)[1], 2))
  facet_labels[i] <- this_facet_label
  
}
#---
df_plot <- as.data.frame(do.call(rbind, list_df))
colnames(df_plot)[1:2] <- c("Time", "Project NPV")
df_plot <- subset(df_plot, s == s[2])
gg <- ggplot(df_plot, aes(Time, `Project NPV`))
#gg <- ggplot(df_cpp, aes(t, cumYt))
gg <- gg + geom_step()
#gg <- gg + facet_wrap(~s, ncol = 1)
gg <- gg + theme_bw()
gg <- gg + theme(axis.text = element_blank(),
                 axis.title.y = element_blank(),
                 axis.title.x = element_text(size = axisTitle_size))
gg_cpp <- gg
dfPlot_cPp <- df_plot
#gg
#---
df_plot <- as.data.frame(do.call(rbind, list_spec))
colnames(df_plot)[1:2] <- c("Logged frequency", "Logged power spectral density")
df_plot <- subset(df_plot, s == s[2])
#names(facet_labels) <- s
gg <- ggplot(df_plot, aes(x = `Logged frequency`, y = `Logged power spectral density`))
gg <- gg + geom_smooth(method = lm, se = F)
gg <- gg + geom_line()
# gg <- gg + facet_wrap(~s, ncol = 1,
#                       labeller = labeller(s = facet_labels))
gg <- gg + theme_bw()
gg <- gg + theme(axis.text = element_blank(),
                 axis.title.y = element_blank(),
                 axis.title.x = element_text(size = axisTitle_size))
gg_cppSpec <- gg

```

```{r Fig1, fig.show = "hold", fig.width = 5, fig.height=2, fig.align="center", fig.cap="\\label{fig:NPVevol}Examples of project NPV evolving over quarterly time steps as (left) a geometric Brownian motion and (right) a compound Poisson process.", echo = FALSE}

gg_gbm + gg_cpp + plot_layout(nrow = 1)

```

In the financial context, the gBm model is eschewed as unrealistic because it does not reflect the abrupt jumps empirically observed in many price series, which are inconsistent with a lognormal size distribution. Instead, "jump" or "jump-diffusion" models are recommended, which more accurately capture these abrupt swings in value [@merton1976option, @tankov2003financial]. An example of a jump process, known as the compound Poisson process (cPp), is displayed in the right panel of Fig \ref{fig:NPVevol} (algorithm 6.2 in Cont and Tankov [-@tankov2003financial]).

The gBm model is methodologically attractive because it allows for a closed form solution to the Black-Scholes PDE. The greater financial realism of jump and jump diffusion models, on the other hand, comes at the high methodological cost of requiring numerical solutions to the Black-Scholes PDE. Corporate ROV methods have adopted the financial context's consensus against gBm in favor of jump and jump diffusion models, thus inheriting the same high degree of methodological complexity. Trigeorgis once characterized this complexity as the unavoidable "bitter pill" that all serious ROV practitioners must come to grips with [-@trigeorgis1993real].

In hindsight, the appetite for bitterness outside of academic audiences---that is to say, adoption of ROV thinking by real world decision makers---has remained low [@horn2015use; @triantis2005realizing; @driouchi2012real; @trigeorgis2018real; @aro2022reflections]. Both critics and proponents alike attribute this tepid reception to the formal complexity of ROV, much of which can be traced back to the use of sophisticated stochastic models and numerical methods [@triantis2005realizing; @aro2022reflections]. From the perspective of research donors and managers, such techniques are effectively black boxes; and even ROV experts find themselves challenged at times. A much cited numerical exercise by Majd and Pindyck [-@majd1987time], for example, was later shown to contain an elementary error [@milne2000time]. It is not always clear, moreover, what is gained in exchange for the bitterness; or whether one set of questionable assumptions is merely replaced with another. The cPp model (e.g., as specified by Pennings and Lint [-@pennings1997option]) rests on the assumption that changes in project NPV occur at random intervals [@tankov2003financial]. Yet substantive reassessments of project NPV typically occur at regular intervals---e.g., through scheduled reviews, milestone evaluations, reporting cycles, or electoral calendars.

This observation about reporting intervals, in turn, points to a deeper consideration that is particularly consequential in the AR4D context: discrete jumps in project NPV are often artifacts of reporting protocols, reflecting not changes in value as they occur, but rather moments when the cumulative effect of a series of such changes is finally made known through a discrete report. This filtration logic applies to both internal management updates and external macro-events. An unexpected government policy shift or field trial result registers as a discontinuous jump only because the continuous, inscrutable processes leading up to it---unobserved ecological dynamics, closed-door political deliberations, gradually shifting coalitions and preferences---remain hidden from view until a formal decree or scheduled measurement and report.

It bears emphasis that this observation extends beyond ecology, where it is relatively obvious, to also encompass social, political, and institutional processes, which are typically experienced as a series of discrete threshold events---collective decisions, policy enactments, norm changes, institutional reforms, and the like. Underlying the apparently discontinuous character of such processes there exists a latent process that may plausibly be treated as continuous. That is to say, the vote is binary, but the deliberation, coalition-building, and preference evolution leading up to it are not; the policy announcement is instantaneous, but the political economy generating it evolves gradually; the institution appears to change abruptly, but the accumulation of contradictions between its formal rules and informal practices, and the gradual realignment of stakeholder interests that precipitate reform are continuous processes operating beneath the surface of apparent stability [@mahoney2009explaining].

It is therefore reasonable to distinguish between a process as reported, which may exhibit discontinuities as an artifact of low-frequency sampling and coarse measurement, and the latent process that is reported, which evolves continuously and independently of the reporting protocol. This continuity becomes still more plausible in the case of aggregate trajectories such as project NPV, which integrate numerous heterogeneous ecological, technical, political, economic, organizational, and social processes.

In the financial and corporate research contexts, this distinction may be largely irrelevant since the publicly reported value of company activities, as reflected in stock prices, is itself the economically relevant object of interest, regardless of what the firm's "true" value may be when insider or otherwise unobservable information is taken into account. In the AR4D setting, however, there is no analogous need to accommodate the filtration structure imposed by measurement and reporting conventions. For purposes of ex-ante impact assessment, the analyst is concerned with the latent stochastic evolution of the project's expected social value itself, regardless of when or how or even if the value is observed and reported. Once this distinction is made, the empirical justification for modeling AR4D project value through jump processes becomes substantially weaker---and the case for a continuous diffusion model correspondingly stronger. Insofar as project NPV reflects the cumulative effect of many small, approximately independent proportional influences of finite variance, functional central limit theorems imply that, as the reporting interval shrinks toward zero, the log-value process may be approximated by Brownian motion, and hence project value by gBm [@oksendal2013stochastic; @billingsley2013convergence].
<!-- Insofar as project NPV reflects the cumulative effect of many small proportional influences that are approximately independent and of finite variance -->
<!-- , and whose fluctuations tend to smooth rather than reinforce one another in the aggregate -->
<!-- that the log-value process converges to Brownian motion in the limit as the reporting interval shrinks to zero, and hence project value to gBm -->
<!-- In other words, there is an important distinction to be made between the discontinuous, measured/reported value trajectory and the latent stochastic value process that is measured and reported. The latent NPV-generating process exists independently of the reporting protocol through which it is periodically observed, and represents an aggregate of numerous research and non-research related stochastic processes---technical, biological, meteorological, political, economic, and institutional---many of which are themselves approximately continuous at fine time scales. It is therefore reasonable to treat the latent value trajectory as continuous for practical modeling purposes. -->
<!-- This is precisely the insight of the gradual institutional change literature [@mahoney2010explaining], which documents how institutions that appear externally stable are in fact continuously evolving through layering, conversion, drift, and displacement---processes that are continuous at the micro level even when their macro-level manifestations appear punctuated. What registers as a discrete institutional discontinuity is typically the public crystallization of a process that has been evolving continuously at the level of interests, coalitions, and informal practices -->
<!-- — a distinction that maps precisely onto the difference between the reported and latent value trajectories developed here. Whether social and institutional processes are better modeled as continuous or discontinuous at the level of their latent micro-dynamics is therefore an open empirical question, and the onus to demonstrate genuine discontinuity in the underlying process — as opposed to in its observation or announcement — lies no less with proponents of jump models than the onus to defend continuity lies with proponents of diffusion models. -->

A separate objection is that AR4D project value may be path dependent---in the sense of David [-@david1985clio] and Arthur [-@arthur1989competing]---whereby historical sequences of events influence future technological and institutional trajectories, a concern that Ruttan [-@ruttan1996induced] discusses in the context of agricultural innovation. While this literature primarily characterizes industry- and system-level dynamics, it raises the concern that project/asset level Markovian valuation models such as gBm may be misspecified if current NPV does not constitute a sufficient summary of past developments, and therefore may fail to capture the effects of mechanisms such as technological lock-in on future value. This concern applies not only to gBm but to all Markovian models of value dynamics, including the jump and jump-diffusion processes just discussed. Nor is it specific to non-market research contexts. The path dependence literature itself originated in analyses of technological adoption in the private sector [@david1985clio; @arthur1989competing], where firm value is likewise shaped by early adoption dynamics, network externalities, and related lock-in mechanisms.

Financial economics has long acknowledged the role of such historical contingencies in shaping fundamentals, and yet asset prices are routinely modeled using Markovian approximations. This is because the canonical efficient markets literature [@samuelson2016proof; @fama1970efficient; @harrison2023martingales] shows that, under rational expectations and sufficiently strong incentives for valuation errors to be corrected, current prices incorporate available information about underlying fundamentals such that past price movements do not provide systematic predictive power. While this martingale characterization does not itself assert that price is a sufficient summary state variable, subsequent practice operationalizes this insight by specifying Markov models of asset value evolution [@black1973valuation; @merton1976option; @hull9thEdition], which effectively treat current price as a sufficient summary---at least for practical decision-making purposes---of the information contained in past developments.

This practical implication---that current value adequately summarizes decision-relevant historical information despite path dependent fundamentals---rests on two complementary premises. The first, formalized in Samuelson's [-@samuelson2016proof] axiom of mathematically expected price formation, holds that if a price were known with certainty to rise tomorrow, it would have already risen today. That is to say, rational agents immediately incorporate any predictable future change into current value, leaving only genuinely new information to drive future revisions. The second, central to Fama's [-@fama1965behavior] empirical argument, is the existence of "sophisticated traders" who continuously scrutinize assets for discrepancies between market price and their own expert assessments of fundamental value, and act to eliminate such discrepancies. Together, these premises enforce the martingale property of current value and underpin the practical adequacy of Markovian approximations. The widespread use of such models therefore reflects not a denial of historical contingency, but a consequence of these rationality conditions, under which the informational consequences of historical contingency are incorporated into current value.

Both of these canonical premises are present in the non-market AR4D setting, if in attenuated form. A rational ex-ante impact assessor of an AR4D project, like Samuelson's price-forming agent, will immediately revise their current NPV estimate upward if they know with certainty that project value will rise. To do otherwise would constitute a straightforward violation of rational expectation formation. And ex-ante impact assessors are, by professional function, precisely analogous to Fama's sophisticated traders: they conduct discrete, independent expert assessments of a project value trajectory shaped by the same class of fundamental processes (economic, political, institutional, environmental, etc.), and revise current estimates whenever their analysis reveals a discrepancy. The essential difference between market and non-market settings therefore lies in the discovery mechanism, not in the nature of the latent value process being discovered. More specifically, the difference is one of sampling rate and resolution. A liquid asset market aggregates large numbers of independent expert assessments per second, washing out individual analyst bias and approximating the martingale ideal at high fidelity. An AR4D impact assessment approximates an analogous latent value, shaped by the same class of historically conditioned fundamentals, at far coarser resolution---periodically rather than continuously, through few analysts rather than many, with weaker incentive-based bias correction.

Even if one grants that substantial discontinuities or path-dependent effects may occasionally arise over the course of an AR4D project's 10-20 year trajectory, the practical significance of this concern is substantially mitigated by the multistage structure of the proposed decision model. The relevant valuation problem is not the direct appraisal of a single uninterrupted multi-decade payoff stream, but rather the sequential valuation of continuation options over comparatively short decision intervals, typically on the order of 1–5 years in the AR4D context. At each stage, updated information is incorporated into revised continuation values, which then serve as the state variables governing subsequent decisions. The operational question is therefore not whether the entire long-run trajectory is perfectly continuous and Markovian, but whether project value may reasonably be approximated as locally continuous and Markovian between adjacent decision nodes after information updating. This is a substantially weaker requirement.
<!-- The multistage framework therefore not only operationalizes staged AR4D decision making, but also narrows the temporal domain over which the diffusion approximation must remain empirically plausible. -->
<!-- An additional advantage of the gBm specification in this regard is that, if project value follows a geometric Brownian motion, the associated continuation option values inherit a locally lognormal and Markovian structure (as shown the next section). -->

Finally, it should be borne in mind that the choice of a stochastic model for project NPV does not constitute a claim that the selected model provides a literal description of reality. In ex-ante AR4D appraisal, the latent evolution of project value is not directly observable, and competing specifications can rarely be ranked through decisive empirical tests. The choice is therefore not between a correct model and incorrect alternatives, but among imperfect approximations that differ in their conceptual plausibility, analytical tractability, transparency to non-specialist users, and usefulness for guiding sequential investment decisions. Evaluated on these criteria, gBm occupies a defensible and arguably privileged position. It is theoretically motivated by functional central limit theorem arguments (as discussed above), consistent with the martingale characterization of rationally updated valuations, free of the information filtration assumptions that burden jump-based alternatives in non-market settings, and sufficiently tractable to remain interpretable to the research donors and managers whose decisions it is ultimately designed to inform.
<!-- Discontinuous jumps appearing in the reported trajectory represent   imposed upon the continuous latent series post hoc, as a function of the reporting artifice by which the latent trajectory becomes observable. -->
<!-- While it is conceivable that discrete jumps in the latent value may occur over the 10-20 year R&D period (due to extreme events such as wars, stock market crashes, serendipitous research externalities, etc.), the benefit of the multi stage approach is that only looking at a 1-5 year horizon at any given decision point. If NPV gBm then any stage ROV is locally lognormal (as shown in the next section)... -->
<!-- the latent value process being estimated has the martingale property independently of the mechanism by which it is assessed. -->
<!-- ; the market simply reveals it at higher resolution. -->
<!-- Markets reveal the true latent value process at higher frequency and fidelity than periodic expert assessment -->
<!-- ---reflecting differences in the mechanism through which true latent value is revealed, not in the nature of the value process being revealed. -->
<!-- , but neither mechanism generates the value process---both merely reveal it. -->
<!-- In the non-market context of AR4D, the no-arbitrage price discovery mechanism does not exist. However, this does not mean that the discovered value does not exist. Trading does not generate the underlying value of an asset, but helps reveal it. It follows that the approximately Markov property of the revealed value process are thus not conferred by the mechanism of revelation, but rather are intrinsic to the process itself and exist independently of it. Again, for purposes of AR4D ex-ante impact assessment, it is this latent value trajectory that is of interest, independent of if, how, when, or to whom it is revealed. The market based finding that the NPV of corporate research activities (as reflected in stock price) adequately approximate a summary state variable therefore transfers to non-market research settings shaped by the same early adoption dynamics, network externalities, and related lock-in mechanisms. -->
<!-- Ok, this provides some good structure, but the argument is this: 1) The market based literature acknowledges that value formation is path dependent/historically conditioned but proves that the current value revealed by markets nonetheless sufficiently summarizes all decision-relevant historical information; and that this is well approximated for practical purposes by Markov models. 2) I argue that, because the market mechanism is revelatory not generative, the market based Markov result may be interpreted as revealing more generally that for any given asset exhibiting path dependence, whether public or private, whether in a market or non-market setting, there exists a value trajectory that, regardless of whether it is ever discovered, for practical valuation purposes sufficiently summarizes all decision-relevant information in its current state, and is thus well approximated by a Markovian model despite its path dependence. 3) Therefore, for a given AR4D project (which by definition occurs in a non-market setting), there exists a latent NPV trajectory, independent of how or when or if it is ever detected or measured, that for practical valuation purposes sufficiently summarizes all decision-relevant information in its current state; and that a Markov approximation of said AR4D value trajectory is thus plausible. -->

<!-- Step 1 — The efficient markets literature establishes empirically and theoretically that asset prices in markets with path dependent fundamentals are adequately approximated by Markovian models for practical valuation purposes. -->
<!-- Step 2 — Because markets are revelatory rather than generative, this result may be interpreted as an existence theorem: it reveals that, for assets with path dependent fundamentals, there exist underlying value processes that are adequately approximated by Markovian models despite their path dependence. This property is intrinsic to the value dynamics themselves, not conferred by the market mechanism. The market demonstrates the existence of such processes; it does not create them. -->
<!-- Step 3 — Ruttan's analysis establishes that the path dependence mechanisms operating in AR4D public good research are of the same character as those documented in corporate technology contexts — the same class of mechanisms for which the market-based existence result was demonstrated. -->
<!-- Step 4 — Therefore, for a given AR4D project, there plausibly exists a latent NPV trajectory that, despite path dependent fundamentals of the same character, is adequately approximated by a Markovian model for practical valuation purposes — independent of whether that trajectory is ever revealed by a market or any other mechanism. -->
<!-- The existence theorem claim belongs in Step 2 because it is a consequence of the revelatory thesis, not a free-standing observation about market practice. The market result in Step 1 is just an empirical regularity. It only becomes an existence theorem once Step 2 establishes that the regularity reflects intrinsic properties of value processes rather than properties of the market mechanism. That is the philosophical move that does the work, and the logical dependence should be explicit. -->

<!-- Moreover, the multistage ROV model proposed below explicitly structures the valuation problem as a sequence of short-horizon decisions. Rather than modeling the full 10–20 year impact trajectory in a single step, the model evaluates a series of continuation values at successive decision nodes (typically 1–5 years apart). Each continuation value is re-estimated using newly available information on technical feasibility, adoption prospects, costs, and expected benefits, thereby embedding the decision-relevant consequences of prior developments since the first decision node into the current state. The Markov assumption is thus invoked locally---between adjacent decision nodes---rather than globally over the entire project lifecycle. Under this formulation, geometric Brownian motion serves as a tractable approximation to the evolution of recursively updated continuation values, not as a literal description of all underlying processes. As shown below, if project value follows a geometric Brownian motion, the associated continuation option values inherit a locally lognormal, Markovian structure, yielding a time-inhomogeneous process with constant drift and time-varying volatility, and preserving internal consistency across stages as uncertainty is updated at each decision node. -->

<!-- However, this concern operates at the level of the structural processes generating value, and does not necessarily imply that a reduced-form valuation process must explicitly track historical paths if current NPV adequately summarizes their decision-relevant consequences. -->

<!-- By the same token, the absence of a market for AR4D benefits does not imply the absence of an underlying project value process that acts as a summary state variable. Rather, it implies that there is no price discovery mechanism facilitating empirical measurement of this value. -->
<!-- Again, the analyst in the non-market context is concerned with the true value trajectory, regarless of whether or not it is ever discovered or reported, nor how, nor when, nor to whom. For the purpose of stochastic model selection in the assessment of project NPV, the abstract existence of such a trajectory is sufficient. -->
<!-- In the non-market context of AR4D, the no-arbitrage price discovery mechanism is absent. This does not imply the absence of an underlying value process, but only the absence of a decentralized mechanism for revealing it. In its place, ex-ante impact assessment serves to approximate this latent trajectory. For decision-making purposes, it is sufficient that such a value process exists and can be represented as a sufficient state variable, regardless of whether it is directly observed in market prices. -->
<!-- OR to close with connection back to the perceived vs. true wording of previous para: Importantly, price discovery is revelatory rather than creative: trading does not generate the underlying value of an asset; it merely reveals and updates information about it. By the same logic, the absence of a market for AR4D benefits does not imply the absence of an underlying, evolving project-value process. It implies only that this true value remains latent—less directly observed and necessarily estimated through appraisal rather than inferred from continuous pricing. Ex-ante impact assessments and periodic project reviews may therefore be understood as discrete, administrative observations of this evolving true value, rather than the moments of its creation. Consequently, if the analytical objective is to model the continuous evolution of a project's true expected value rather than the discrete jumps of its perceived, reported value, the system can still be rigorously represented through a reduced-form Markovian approximation for sequential decision making. -->
<!-- The central issue, then, is whether AR4D project value can be represented as an underlying state process whose current level sufficiently summarizes the decision-relevant consequences of prior developments (and thus a path independent process). This latent valuation process need not be directly observable. In practice, ex-ante impact assessments and project reviews provide only discrete, filtered observations shaped by institutional reporting protocols. Different protocols may therefore generate different observed value paths. For modeling purposes, however, it is analytically useful to consider the limiting idealization in which such reporting frictions are abstracted away, yielding a continuously evolving project-value process to which reduced-form stochastic approximations such as gBm may be applied. Such an approximation is not implausible in the present context, since project NPV is itself a synthesis of numerous underlying drivers—biophysical, behavioral, institutional, and macroeconomic—many of which evolve incrementally over time, so that their combined effect may reasonably be represented as a smoother continuous trajectory than any single component viewed in isolation. -->
<!-- ...Ex-ante impact assessments and periodic project reviews may therefore be understood as discrete observations of this evolving latent state variable. As such, project value can still be rigorously represented through a reduced-form Markovian approximation for sequential decision making. -->

<!-- The specific choice to model this latent Markovian process as a geometric Brownian motion (gBm) is driven by the structural realities of agricultural welfare generation. First, the Markov property asserts that the current state variable acts as a sufficient statistic; it does not deny historical path dependence, but rather demands that the economic consequences of that history are fully absorbed into the current valuation. Second, the fundamental drivers of AR4D value—biophysical realities, agronomic outcomes, and decentralized adoption decisions—manifest as a multitude of independent, proportional shocks. Under standard stochastic limit theorems, the aggregate mathematical limit of these multiplicative micro-events converges to a continuous gBm. Finally, in the data-sparse environment of ex-ante appraisal, gBm serves as the maximum-entropy baseline. It enforces the necessary boundary condition of strictly positive value while avoiding the severe, weakly identified parameter burdens inherent to jump-diffusion or mean-reverting alternatives. Therefore, gBm is not deployed as a perfect ontological reflection of agricultural innovation, but as the most parsimonious, theoretically justified benchmark available to the practitioner. -->
<!-- Finally, this distinction clarifies the precise target of ex-ante impact assessment and brings the theoretical argument against jump processes full circle. As established earlier, because a project's true expected value is a synthesis of many continuous component processes—biological growth, decentralized social adoption, and gradual political momentum—the latent value itself evolves continuously. The perceived "jumpiness" that traditionally motivates jump-diffusion models is merely an artifact of artificial, discrete information filtration. While there exist infinitely many valid, yet inscrutable, discrete trajectories for a project's reported NPV—each corresponding to a different institutional reporting protocol (e.g., monthly updates, annual reviews, decadal audits)—the analyst is not interested in modeling the arbitrary friction of a specific administrative schedule. Rather, the analytical objective is to capture the trajectory in the mathematical limit as all artificial filtrations are removed. At this continuous limit, the unfiltered flow of proportional information naturally converges to a diffusion process, firmly establishing gBm not as a compromise, but as the most theoretically consistent and transparent foundational benchmark. -->

<!-- The argument has essentially 3 points: -->

<!-- 1) Many determinants of value are path dependent in both market and non-market settings. Yet valuation models often employ Markovian approximations when current value is treated as a sufficient summary of the economically relevant consequences of that history. -->

<!-- 2) In market contexts, decentralized trading provides a mechanism through which information relevant to value is aggregated into prices. The market does not create the underlying value consequences themselves; it provides an observable signal of them. -->

<!-- 3) In non-market AR4D settings, the absence of market prices affects observability rather than the conceptual existence of an underlying project value process. Ex-ante impact assessments and periodic project reviews provide partial measurements of that evolving value, analogous to discrete observations of a latent state variable. -->

<!-- And then finally this line of argumentation: -->
<!-- A further consideration is the staged structure of many AR4D programs. Although ultimate impact horizons may extend 10–20 years, major continuation, redesign, scaling, or termination decisions commonly occur at intervals of 1–5 years. In a multistage real-options framework, the object valued at an early stage is therefore not the distant terminal payoff in raw form, but an option to proceed to the next stage at the forthcoming decision node. The relevant valuation horizon for the current decision is thus the life of that intermediate option, not the full multi-decade impact horizon. Moreover, if the underlying multistage project NPV is represented as a gBm, then the recursively defined continuation option values inherit the same tractable stochastic form (as shown in the next section). Long-run health, nutrition, environmental, and income consequences therefore enter current value indirectly through these recursively updated continuation values. The practical Markov assumption applies primarily to the evolution of shorter-horizon continuation values between adjacent decision nodes, rather than to the entire 20-year development trajectory viewed as a single uninterrupted process. -->
<!-- [END DEFENSE OF GBM] -->
<!-- the particular administrative or institutional schedule according to which information about that value becomes observable. -->
<!-- , under which the grounds for swallowing the bitter pill of jump and jump-diffusion models are particularly narrow in the AR4D context: -->
<!-- This observation---that discrete jumps in project NPV are often artifacts of reporting protocols---implies a distinction between reported project value and the latent stochastic value process that those reports periodically reveal. Under this interpretation, the grounds for swallowing the methodological “bitter pill” of jump and jump-diffusion models become particularly narrow in the AR4D context. -->
<!-- The distinction between reported project value and the latent stochastic value process that those reports periodically reveal implicit in observation points to a deeper problem with the application of jump and jump-diffusion models in the AR4D context. The apparent discontinuities observed in project NPV time series may arise less from genuinely discontinuous changes in underlying project value than from the institutional structure of information revelation itself. ---that evolve continuously through time even when information about them is disclosed only intermittently. -->
<!-- Which brings up possibly an even more important point regarding why jump/jump diffusion models are a bad fit in the AR4D context: The discrete jump character of project NPV time series is an artifact of reporting protocols. The "true" NPV data generating process that is filtered exists independently of the reporting protocol and is itself a function of numerous research and non-research related stochastic processes that are intrinsically continuous---the direction of future demand, other price-based marketing considerations, the technical details of the research (in the case of agricultural research this includes) biological, meteorological, political, etc.---and is thus itself intrinsically continuous. This makes sense in the corporate context, where the _perceived_ or publicly disclosed value of company activities based on publicly reported information (i.e., its stock price) is what counts as a measure of value, regardless of what the company's _true_ value may be when insider or otherwise inscrutable information is taken into account. In the AR4D setting, however, there is no need to accommodate these artificial reporting filtrations. For the purpose of ex-ante evaluation of future project impacts, the analyst is concerned solely with the intrinsic stochastic evolution of the project's true expected value, regardless of if, when, how, or to whom this information is publicized. Once this distinction is made, the empirical case for modeling value through jump processes becomes less clear-cut. To the extent that project NPV reflects the cumulative effect of many small proportional influences that are approximately independent and of finite variance, functional central limit theorems imply that, abstracting from discrete reporting filtrations, the log-value process may be approximated by Brownian motion, and hence project value by geometric Brownian motion [@oksendal2013stochastic; @billingsley2013convergence]. -->
<!-- The adoption of jump models makes sense in the corporate context, where the _perceived_ value of company activities based on publicly reported information (i.e., its stock price) is what counts as a measure of value, regardless of what the company's _true_ value may be when insider or otherwise inscrutable information is taken into account. In the AR4D setting, on the other hand, there is no need to accommodate these artificial reporting filtrations. For the purpose of ex-ante evaluation of future project impacts, the analyst is concerned solely with the abstract, stochastic evolution of the project's true expected value as information is generated, not when it is publicized. Once this distinction is made, the empirical case for modeling value through jump processes becomes less clear-cut. To the extent that project NPV reflects the cumulative effect of many small proportional influences that are approximately independent and of finite variance, functional central limit theorems imply that, abstracting from discrete reporting filtrations, the log-value process may be approximated by Brownian motion, and hence project value by geometric Brownian motion [@oksendal2013stochastic; @billingsley2013convergence]. -->
<!-- To the extent that NPV reflects the cumulative effect of many small proportional influences that are approximately independent and of bounded variance, stochastic limit theorems imply that as the discrete filtrations of both internal and external reporting are lifted, the mathematical limit of this continuous flow of information converges to a gBm. -->
<!-- Assuming these underlying micro-events act as proportional shocks to project value and are approximately independent and of finite variance,  -->

<!-- typically interested in the latent evolution of true NPV regardless of the timing of administrative or public announcements that determine if, when, how, and to whom this evolution is publicized.  standard diffusion approximations support modeling latent value dynamics with a process such as gBm. -->
<!-- OR -->
<!-- In the non-market AR4D setting, , regardless of whether the withholding entity is the research organization itself or an external government agency. The practitioner   Thus, the high methodological cost of jump mechanics is only justified if one is forced to model the discrete public release of information, rather than the continuous generation of the information itself. -->

<!-- so much for the jump models... -->

<!-- path dependence -- any relevant historical effects are either small, temporary, already embedded in current project valuation, or second-order relative to other uncertainties. -->
<!-- A more serious objection that arises in the AR4D context that applies to gBm and cPp alike (and all models of the Markov class) is that of path dependence... -->
<!-- We acknowledge that many underlying determinants of AR4D impacts—such as weather sequences, soil dynamics, technology adoption, and institutional change—may exhibit path dependence [@ruttan1996induced]. However, project NPV is not identical to any one of these processes. Rather, it is a forward-looking valuation of expected public-good benefit streams, including gains in health, nutrition, environmental services, and smallholder incomes. In this aggregate form, project NPV summarizes the current welfare implications of past developments rather than inheriting each underlying historical dependency in full. Consequently, path dependence in component processes does not automatically imply that the project valuation process itself must be strongly path dependent. A similar logic underlies many financial valuation models, in which valuation variables are often represented using Markovian stochastic approximations because current values summarize the present implications of historically conditioned fundamentals. -->

<!-- Furthermore, the benefit streams underlying AR4D project NPV are realized through the decentralized decisions of many independent agents across heterogeneous contexts, and often do not exhibit, at least at the project valuation margin, the network externalities, increasing returns, or switching costs that are canonical sources of strong system-level lock-in. Thus, while path dependence may characterize some sectoral processes, its transmission to project-level valuation dynamics is not automatic and may be attenuated in aggregate. -->

<!-- This does not imply that all AR4D projects literally follow gBm, nor that genuine discontinuities never occur. Rather, it suggests that the added complexity of jump-based models is justified when discrete shocks, regime changes, or reporting protocol are central to the decision problem. Where such features are secondary or weakly evidenced, gBm remains a transparent and tractable benchmark. -->
<!-- For purposes of ex-ante assessment of project NPV in non-market settings, on the other hand, the stochastic evolution of a project's true value is the time series of interest, regardless of when or even if changes in true NPV are ever registered. -->
<!-- No stochastic specification is likely to describe AR4D project value perfectly. The relevant question is therefore comparative rather than absolute: which simplifying assumptions are most defensible for the decision at hand? While gBm imposes familiar restrictions such as proportional shocks and path-independent increments, jump models impose restrictions of their own. In the compound Poisson case, value revisions are assumed to arrive at random intervals generated by a Poisson process. Yet in AR4D settings, substantive reassessments of project NPV commonly occur through scheduled reviews, milestone evaluations, reporting cycles, or electoral calendars. Thus, the discrete appearance of valuation changes does not by itself support a Poisson jump specification; it may instead reflect institutional timing conventions. Under such conditions, gBm may remain a more transparent benchmark than a richer but differently misspecified jump process. -->

<!-- As mentioned above, it is not even clear what a realistic time evolution of AR4D project NPV looks like. The time and expense involved in AR4D ex-ante impact assessment make frequent, regular updates of project NPV impractical, which in turn means that direct empirical inspection of the stochastic character of project NPV over a series of small, regularly spaced time steps is generally not possible---certainly not in the way that is possible for a price series. -->

<!-- In the corporate context, Pennings and Lint [-@pennings1997option] attempt to resolve this issue by assembling from scratch an historical time series of project NPV; and then selecting and parameterizing a cPp model of future project NPV based on this. Such an approach does not so much resolve the issue as it does replace it with another set of problems. First and foremost, the historical project NPV must be carefully selected so as to be representative of the NPV trajectory of the new project under evaluation. Apart from a passing verbal assurance, Pennings and Lint offer little indication that their assembled time series may be plausibly construed as representative of the stochastic character of the new project under evaluation [-@pennings1997option]. In the AR4D context, especially, projects are so heterogeneous in their scale, scope, aims, target populations and environments, and so forth, that finding a past project NPV trajectory that plausibly maps onto a present project's NPV amounts to a prohibitive challenge. -->

<!-- Secondly, even if this were possible, the practitioner then runs into small sample problems. In both the corporate and AR4D contexts, the number of substantive adjustments to project NPV over the project's lifespan tends to be small. To generate "enough" data points, Pennings and Lint merge several past project NPV trajectories together into a single time series. In addition to severely compounding the first problem, the authors come up with just 17 observations over a 5 year period. Is this enough to select and parameterize a stochastic model of project NPV? -->

<!-- Thirdly, the cPp model selected by Pennings and Lint is appropriate only insofar as the interval size between changes in NPV is random [@tankov2003financial]. This is problematic since substantive changes in project NPV can often occur as the result of non-randomly spaced events, e.g. at scheduled test points in the research cycle, or scheduled shifts in the enabling environment (such as elections). Moreover, such changes are typically documented and announced through periodic, non-random reporting protocols. -->

<!-- Finally, there is the subtle and somewhat philosophical sounding, but deeply consequential, issue of sampling rate. When does a change in project NPV occur? Does it occur when the relevant causal event or set of events occurs, or only when management learns of its occurrence and announces the corresponding NPV adjustment? What about events that effectively change NPV, but that go unnoticed by management? Pennings and Lint tacitly presume that changes in NPV occur only if and when management learns of their occurrence. This is perhaps understandable in the corporate context, where the _perceived_ value of a company based on publicly reported information (i.e., its stock price) is what counts as a measure of value, regardless of what the company's _true_ value may be when insider or otherwise inscrutable information is taken into account. For purposes of ex-ante assessment of project NPV in the non-market AR4D setting, on the other hand, the stochastic evolution of a project's true value is the time series of interest, regardless of when or even if changes in NPV are ever discovered and documented. -->

<!-- This focus on true NPV significantly simplifies ex-ante modeling in non-market settings. In corporate contexts, the reliance on jump and jump-diffusion models---the "bitter pill" of methodological complexity---arises largely because the target variable is perceived value. Modeling perceived value requires explicitly accounting for the information filtration process; the model must capture not just the continuous generation of information, but the discrete administrative mechanisms (e.g., scheduled testing, quarterly earnings reports) through which it is formally released to the market. Perceived value remains static between these scheduled releases, necessitating a jump process to model the sudden market reaction when new information is finally announced. -->

<!-- In the far-from-market AR4D setting, however, there is no need to accommodate these administrative artifacts. The practitioner is concerned solely with the abstract, stochastic evolution of the project's true expected value. As underlying biophysical, social, and institutional realities continuously shift, the true expected value of the project evolves in a mathematically smooth trajectory, regardless of whether or when these shifts are formally measured, reported, or known. The high methodological cost of jump mechanics is only justified if one is forced to model the discrete release of information, rather than the continuous, fluid generation of the information itself. -->
<!-- OR -->
<!-- Furthermore, what appears as a discrete jump in a project's historical NPV is often merely an artifact of blunt measurement---the moment a continuous shift in expected value is finally documented in a periodic project report. In the far-from-market AR4D setting, there is no need to accommodate these administrative artifacts. The practitioner is concerned solely with the abstract, stochastic evolution of the project's true expected value. As underlying biophysical, social, and institutional realities continuously shift, the true expected value of the project evolves, regardless of whether or when these shifts are formally reported. Indeed, assuming these underlying micro-events act as proportional shocks to project value, stochastic limit theorems dictate that as the artificial discrete filtrations of administrative reporting are lifted, the mathematical limit of this continuous flow of information converges strictly to a geometric Brownian motion (gBm). Thus, the high methodological cost of jump mechanics is only justified if one is forced to model the discrete release of information, rather than the continuous generation of the information itself. -->

<!-- This raises a fundamental distinction between the discrete realization of an innovation event and the continuous evolution of its expected value. It is undeniable that agricultural innovation processes—particularly those involving institutional transformations, policy shifts, or collective action—often culminate in discrete, path-dependent events. If project NPV were merely a retrospective ledger of these realized events, a jump process might be appropriate. -->

<!-- Yet, ex-ante NPV is inherently a measure of expected value. While the final realization of a policy decree or a community consensus is a discrete occurrence, the social, political, and biophysical deliberations leading up to that shift are fluid. As information regarding the likelihood of these events continuously emerges in the target environment, the probability of success—and thus the expected NPV of the project—evolves in a mathematically smooth trajectory. Modeling project NPV as a jump process falsely assumes that the probability of a successful outcome remains completely static until the moment an event occurs. In reality, the inscrutable data generating process of project NPV reflects this continuous flow of information and shifting expectations. -->

<!-- Furthermore, what appears as a discrete jump in a project's historical NPV is often merely an artifact of blunt measurement---the moment a continuous shift in expected value is finally documented in a periodic project report. Therefore, modeling the time evolution of AR4D project NPV as a continuous geometric Brownian motion (gBm) is not merely an assumption of methodological expedience. It is a theoretically rigorous reflection of how the expected value of path-dependent agricultural transformations actually evolves prior to their formal documentation. This approach preserves the functional integrity of the option valuation while purposefully abstracting away the intractable mathematical overhead of jump-diffusion mechanics. -->

<!-- In non-market contexts such as AR4D, on the other hand, one is at liberty to concern themselves only with the stochastic progression of real changes to project NPV, regardless of when management, or anyone else, discovers and documents these changes. -->
<!-- As in the corporate examined by Pennings and Lint, research related changes in NPV predominate the -->

<!-- For present purposes, then, the periods of no change in the cPp model are effectively equivalent to periods of negligibly small changes in the gBm model; and the size distribution of changes in NPV generated by the two models is thus also effectively the same. The choice between gBm and cPp thus boils down to a choice between a [[reasonably realistic]--here's what I have to justify], methodologically expedient model, and a model that may or may not slightly improve realism at the cost of a large methodological headache. In the present work, the expedient model (gBm) is chosen. -->

<!-- Conditions under which this is valid -->

<!-- Finally, there is the subtle and somewhat philosophical sounding, but deeply consequential, issue of sampling rate. When does a change in project NPV occur? Does it occur when the relevant causal event or set of events occurs, or only when management learns of its occurrence and announces the corresponding NPV adjustment? Pennings and Lint tacitly presume the latter. This is perhaps understandable in the corporate context, where the value of company activities is ultimately defined by the stock market, and hence by publicly available information. In non-market contexts such as AR4D, however, the underlying drivers of project NPV—incremental biophysical realities, climatic shifts, and evolving socio-economic conditions—do not freeze in place between scheduled reports. Assuming that project NPV remains perfectly flat for months and then instantaneously jumps to a new value mistakes the blunt artifact of an administrative reporting cycle for the underlying reality. -->

<!-- In reality, the net change documented in any given quarterly or annual evaluation is the aggregate sum of countless unobservable, intermediate micro-events. While agricultural innovation processes are frequently discontinuous and path-dependent at the micro-level, stochastic theory provides a formal bridge to macroscopic modeling. Specifically, the Functional Central Limit Theorem (Donsker's Invariance Principle) demonstrates that the accumulation of independent, discrete shocks—regardless of their underlying distribution or discontinuity—converges toward a continuous Brownian motion when aggregated over macroscopic time intervals. Therefore, modeling the quarter-to-quarter progression of project NPV as a continuous geometric Brownian motion (gBm) is not an oversimplification. Rather, it provides a mathematically rigorous approximation of how unobserved variance naturally aggregates over administrative time steps. This approach preserves the functional integrity of the option valuation while purposefully abstracting away the intractable mathematical overhead of jump-diffusion mechanics. -->

<!-- At the resolution of a quarterly or annual evaluation cycle, the precise sequence of the continuous "buzz" and the exact timing of any discrete research "jump" blur together into a single, net aggregate change. According to the multiplicative version of the Central Limit Theorem, the accumulation of the constant background perturbations naturally converges toward a lognormal distribution. When the occasional, larger research-related shock is absorbed into this discrete time step, the resulting net percentage change in project NPV over that quarter is mathematically well-behaved and can be reasonably mapped to a lognormal probability density function. -->

<!-- Therefore, while the underlying, micro-level mechanics of agricultural innovation are indeed discontinuous and path-dependent, modeling the resulting macro-level, quarter-to-quarter net variance as a gBm is not an oversimplification; it is a pragmatic recognition of how variance aggregates over time. It provides a logically defensible fit for the practical data-generating process, purposefully avoiding the intractable mathematical overhead of jump-diffusion mechanics that have historically alienated decision-makers from real option valuation. -->

<!-- Because AR4D project NPV time series do not exist, the choice of size distribution of changes in NPV requires deductive reasoning. A reasonable starting point in this exercise is the observation that changes in project NPV seem to be of two types: research related and non-research related. -->

<!-- Research related changes in project value generally occur at discrete test points, when new information regarding the effectiveness of the new technology becomes available. Given the characteristic pace of AR4D experimentation, such changes by definition occur infrequently, perhaps 1-4 times per year, and are typically significant in size. These large research related shifts occur, meanwhile, against the backdrop of a more or less constant "buzz" of non-research related perturbations to NPV---mostly undocumented or too small to register in project reports, but nonetheless real---reflecting developments of a political, socio-economic, institutional, and financial nature in the target populations and production environments where the new technology is to be released. -->

<!-- The size distribution of changes in AR4D project NPV may thus be reasonably characterized as one that is predominated by many frequent small changes, punctuated by infrequent large changes.  -->
<!-- Such changes may include, for example, elections, policy changes, commodity price swings, changes in seed systems and other value chain mechanisms, changes in the security environment, the ebb and flow of public and private sector partnerships to enhance impact, and so forth. -->
<!-- Research related changes occur relatively infrequently, perhaps 1-4 times per year, but with relatively greater magnitude than non-research related changes which occur more frequently, perhaps 1-4 times per annual quarter. Overall, then, it is reasonable to expect changes in AR4D project NPV to occur in every quarter, but that most of these changes will be small, even negligibly small, interspersed by a few larger, substantive changes. Based on this rationale, the lognormal distribution is a reasonable choice as a model of the size distribution of percentage changes in project NPV per annual quarter. The corresponding stochastic process is the gBm, an example of which is depicted in the left panel of Fig \ref{fig:NPVevol}. -->

# Formal derivation of multistage ROV in the AR4D context

## Derivation of single stage ROV \label{sec:straightInt}

Assuming $x(t)$ follows a gBm, the evolution $\Delta x$ over a small time increment $\Delta t$ is described by

\begin{equation}
\Delta x = x(t) m \Delta t + x(t) s \epsilon \sqrt{\Delta t} \:\:;\:\:\: \Delta x := x(t + \Delta t) - x(t)
\label{eq:gbmEq}
\end{equation}

Where $\epsilon$ is a normally distributed random variable with mean $0$ and variance $1$, such that $m$ and $s$ quantify the trend and volatility of the time evolution, respectively.

\begin{equation}
\begin{split}
m \Delta t &= E \left[\frac{\Delta x}{x} \right] \\
s^2 \Delta t &= Var \left[\frac{\Delta x}{x} \right]
\end{split}
\end{equation}

In other words, $m \Delta t$ and $s^2 \Delta t$ are the mean and variance of the instantaneous arithmetic return or growth rate. The expected value of project NPV at some future time $T$ as evaluated at time $t = 0$ is given by

\begin{equation}
E[x(T)]\bigr|_{t = 0} = e^{m T} x(0)
\label{eq:ExTt}
\end{equation}

And the mean and variance of the log return over any finite interval $[0, T]$ are given by

\begin{equation}
E \left[ \ln \left(\frac{x(T)}{x(0)} \right) \right]\biggr|_{t = 0}  = \left(m - \frac{s^2}{2} \right) T
\label{eq:meanLn}
\end{equation}

\begin{equation}
Var \left[ \ln \left(\frac{x(T)}{x(0)} \right) \right]\biggr|_{t = 0} = s^2 T
\end{equation}

(See Hull [-@hull9thEdition] for details.)

Since the calculation of project NPV already accounts for all expected changes to NPV that might occur over the time horizon, then, by definition, the expected growth rate of project NPV is just the discount rate ($m = r$).

The single stage ROV formula follows by evaluating Eq \eqref{eq:basic} through straightforward integration.

In general, for a lognormally distributed random variable $q$ and a constant $C$,

\begin{equation}
E[\max(q - C, 0)] = E[q] \Phi \left(\frac{\ln \left( \frac{E[q]}{C} \right) + \frac{\omega^2}{2}}{\omega} \right) - C \Phi \left(-\frac{\ln \left( \frac{E[q]}{C} \right) - \frac{\omega^2}{2}}{\omega} \right)
\label{eq:genEq}
\end{equation}

Where $\Phi()$ is the standard cumulative normal distribution function. (See S1 Appendix for proof.)

In the case where $q = x(T_0)$, $C = K_0$, and multiplying through by the discount factor $e^{-r T_0}$, this becomes

\begin{equation}
e^{-r T_0} E[\max(x(T_0) - K_0, 0)]\bigr|_{t = 0} = x(0) \Phi \left(\delta_0 + s \sqrt{T_0} \right) - e^{-r T_0} K_0 \Phi(\delta_0)
\label{eq:rov}
\end{equation}

Where

\begin{equation}
\delta_0 = \frac{ \ln \left(\frac{x(0)}{K_0} \right) + \left(r - \frac{s^2}{2} \right) T_0} {s \sqrt{T_0}}
\end{equation}

Note that the value of $\Phi(\delta_0)$ is instructive. It is the probability that $x(T_0)$ will exceed the exercise cost $K_0$---or, in real options parlance, the probability that the option finishes "in the money". To see this, note that $\delta_0$ may be rewritten as the negative standard score of $\ln(K_0/x(0))$.

\begin{equation}
\begin{split}
\delta_0 &= -\frac{ \ln \left(\frac{K_0}{x(0)} \right) - \left(r - \frac{s^2}{2} \right) T_0} {s \sqrt{T_0}} \\
&= -\frac{ \ln \left(\frac{K_0}{x(0)} \right) - E \left[ \ln \left(\frac{x(T_0)}{x(0)} \right) \right]\bigr|_{t = 0}} {s \sqrt{T_0}}
\end{split}
\end{equation}

Hence,

\begin{equation}
\begin{split}
\Phi(\delta_0) &= 1 - \Phi \left( \frac{ \ln \left(\frac{K_0}{x(0)} \right) - E \left[ \ln \left(\frac{x(T_0)}{x(0)} \right) \right]\bigr|_{t = 0}} {s \sqrt{T_0}} \right) \\
&= 1 - P \left( \ln (x(T_0)) \leq \ln ( K_0)\right) \\ &= P \left( \ln (x(T_0)) \geq \ln (K_0) \right)
\end{split}
\end{equation}

## Derivation via the Itô calculus and relation to the Black-Scholes derivation

Geometric Brownian motion is a particular case of a broader class of stochastic processes known as Itô processes, defined as follows.

\begin{equation}
\Delta x = a(x, t) \Delta t + b(x, t) \epsilon \sqrt{\Delta t}
\end{equation}

Where $a$ and $b$ are functions of $x$ and $t$.
<!-- In the present case where $x(t)$ is project NPV, $a = x(t) r$ -->

Itô's lemma states that the evolution of a function $f(x, t)$ is also an Itô process, described as follows [@hull9thEdition].

\begin{equation}
\Delta f = \left(\frac{\partial f}{\partial x} a + \frac{\partial f}{\partial t} + \frac{1}{2} \frac{\partial^2 f}{\partial x^2} b^2 \right) \Delta t + \frac{\partial f}{\partial x} b \epsilon \sqrt{\Delta t}
\end{equation}

In the present case, where $x(t)$ is a gBm with $a = x r$, $b = x s$, Itô's lemma reduces to

\begin{equation}
\Delta f = \left( \frac{\partial f}{\partial x} x r + \frac{\partial f}{\partial t} + \frac{s^2 x^2}{2} \frac{\partial^2 f}{\partial x^2} \right) \Delta t + \frac{\partial f}{\partial x} x s \epsilon \sqrt{\Delta t}
\label{eq:itoLem}
\end{equation}

It follows that the expected value and variance of the instantaneous growth rate of $f$ are, respectively,

\begin{equation}
E\left[ \frac{\Delta f}{f} \right] = \frac{1}{f} \left( \frac{\partial f}{\partial x} x r + \frac{\partial f}{\partial t} + \frac{s^2 x^2}{2} \frac{\partial^2 f}{\partial x^2} \right) \Delta t
\label{eq:Ef1}
\end{equation}

and

\begin{equation}
Var\left[ \frac{\Delta f}{f} \right] = \eta_{1,0}^2 s^2 \Delta t
\end{equation}

Where the shorthand $\eta_{1,0} = \frac{x}{f} \frac{\partial f}{\partial x}$ has been introduced. (See Hull [-@hull9thEdition] for details.)

The function $f(x, t)$ in mind here is, of course, ROV (Eq \eqref{eq:rov}). Because ROV is defined as an expected future value discounted back to the present, then its expected instantaneous growth rate is, by definition, $E[\Delta f / f] = r \Delta t$. Eq \eqref{eq:Ef1} may thus be set equal to $r \Delta t$, resulting in the following PDE.

\begin{equation}
fr = \frac{\partial f}{\partial x} x r + \frac{\partial f}{\partial t} + \frac{s^2 x^2}{2} \frac{\partial^2 f}{\partial x^2}
\label{eq:bsPDEar4d}
\end{equation}

This may then be solved for $f$ under the boundary condition $f(T_0) = \max(x(T_0) - K_0, 0)$, resulting in the same expression obtained through straightforward integration in Eq \eqref{eq:rov}. (See S1 Appendix for details.)

To reach the same result in the financial context, Black and Scholes [-@black1973valuation] famously noted that Eqs \eqref{eq:gbmEq} and \ref{eq:itoLem} could be combined so as to eliminate the random term as follows.

\begin{equation}
\Delta f - \frac{\partial f}{\partial x} \Delta x = \left(\frac{\partial f}{\partial t} + \frac{s^2 x^2}{2} \frac{\partial^2 f}{\partial x^2} \right) \Delta t
\label{eq:bsInsight}
\end{equation}

In the financial context, the left-hand side of this equation may be thought of as the instantaneous evolution over the increment $\Delta t$ of a portfolio long one share of the financial derivative $f$ and short a quantity $\partial f / \partial x$ of the underlying security $x(t)$. This is where Black and Scholes applied their no-arbitrage argument: Since the random---i.e. risky---term has been eliminated from Eq \eqref{eq:bsInsight}, then the profit or loss of this portfolio over the increment $\Delta t$ must be riskless. That is, it must change at the risk free rate $r$.

\begin{equation}
\Delta f - \frac{\partial f}{\partial x} \Delta x = \left( f - \frac{\partial f}{\partial x} x \right) r \Delta t
\label{eq:BSarb}
\end{equation}

Equating the right-hand side of this equation with the right-hand side of the previous equation, the $\Delta t$'s cancel, resulting in Eq \eqref{eq:bsPDEar4d}.

Note that the Black-Scholes derivation hinges critically upon the assumption that $x(t)$ follows a gBm. As discussed in the previous section, this assumption is largely rejected in the literature subsequent to Black and Scholes' seminal paper, thus rendering the closed form solution in Eq \eqref{eq:rov} of mere academic interest in the financial and corporate ROV contexts.
<!-- \ref{sec:gBmDefens} -->

To tidy up, note that, since the expected growth rate of ROV is defined as $r\Delta t$, then the time evolution of $f$ in Eq \eqref{eq:itoLem} may be rewritten more simply as follows.

\begin{equation}
\Delta f = f r \Delta t + f \eta_{1,0} s \epsilon \sqrt{\Delta t}
\end{equation}

A further consequence of Itô's lemma is that log changes in $f$ are normally distributed with mean and variance

\begin{equation}
E\left[\ln \left( \frac{f(T_1)}{f(0)} \right) \right]\biggr|_{t = 0} = \left(r - \eta_{1,0}^2 \frac{s^2}{2} \right) T_1
\label{eq:muLf1_b}
\end{equation}

\begin{equation}
Var\left[\ln \left( \frac{f(T_1)}{f(0)} \right) \right]\biggr|_{t = 0} = \eta_{1,0}^2 s^2 T_1
\label{eq:sigLf1}
\end{equation}

While the Itô calculus is not necessary for deriving the single stage ROV formula (this was demonstrated in the previous subsection), it becomes useful in the derivation of a formula for multistage ROV because it provides a clear demonstration that ROV itself is lognormally distributed with computable parameters given in Eqs \eqref{eq:muLf1_b} and \eqref{eq:sigLf1}; and that this is recursively true for the ROV of $f$, and for the ROV of the ROV of $f$, and so on, ad infinitum, as detailed in the next section.
<!-- \ref{sec:straightInt} -->

## Extension to multistage ROV \label{sec:method}

For the derivation of a multistage ROV formula, it is henceforth necessary to index $f$ as $f_1$, $f_2$,..., on up to $f_n$. (The $f_i$ correspond to the $ROV_i$ appearing in the Introduction). Following this schema, $f_1$ corresponds to Eq \eqref{eq:rov} and to $f$ in the equations above, and $f_0$ corresponds to project NPV $x$.

Because $f_1$ is itself lognormally distributed, the same formula used to evaluate $f_1$ can be used to evaluate $f_2(T_2)$. That is to say, Eq \eqref{eq:genEq} may be evaluated at $q = f_1(T_1)$ and $C = K_1$ to calculate $f_2$, the value of the option to invest in the last stage of research.

\begin{equation}
\begin{split}
f_2(f_1, t; T_1, K_1, s, r)|_{t = 0} &= e^{-r T_1} E[\max(f(T_1) - K_1, 0)]\bigr|_{t = 0} \\
&= f_1(0) \Phi(\delta_1 + \eta_{1,0} s \sqrt{T_1}) - e^{-r T_1} K_1 \Phi(\delta_1)
\end{split}
\label{eq:rov2stage}
\end{equation}

Where $T_1$ is the time until completion of stage 1, $K_1$ is the research investment required upon successful completion of stage 1---where stage 1 is considered successfully completed if $f_1(T_1) > K_1$---and where

\begin{equation}
\delta_1 = \frac{ \ln \left(\frac{f_1(0)}{K_1} \right) - \left(r - \eta_{1,0}^2 \frac{s^2}{2} \right) T_1} {\eta_{1,0} s \sqrt{T_1}}
\end{equation}

Moreover, because $f_1$ is an Itô process and $f_2$ is a function of $f_1$, then by Itô's lemma, $f_2$ is also an Itô process,

\begin{equation}
\Delta f_2 = \left( \frac{\partial f_2}{\partial f_1} r f_1 + \frac{\partial f_2}{\partial t} + \eta_{1, 0}^2 \frac{s^2 f_1^2}{2} \frac{\partial^2 f_2}{\partial f_1^2} \right) \Delta t + f_2 \eta_{2, 0} s \epsilon \sqrt{\Delta t}
\label{eq:itoLem2}
\end{equation}

With expected value and variance of the instantaneous growth rate of $f_2$

\begin{equation}
E\left[ \frac{\Delta f_2}{f_2} \right] = \frac{1}{f_2} \left( \frac{\partial f_2}{\partial f_1} r f_1 + \frac{\partial f_2}{\partial t} + \eta_{1, 0}^2 \frac{s^2 f_1^2}{2} \frac{\partial^2 f_2}{\partial f_1^2} \right) \Delta t
\label{eq:Ef2}
\end{equation}

\begin{equation}
Var\left[ \frac{\Delta f_2}{f_2} \right] = \eta_{2,0}^2 s^2 \Delta t \:\:;\:\:\: \eta_{2,0} = \frac{f_1}{f_2} \frac{\partial f_2}{\partial f_1} \eta_{1, 0}
\end{equation}

And such that the log return $\ln(f_2(T)/f_2(t))$ for some future time $T \in (0, T_1)$ is normally distributed with expected value and variance
<!-- Where $T_2$ is the time until completion of stage 1 of the 2-stage project.  -->

\begin{equation}
\begin{split}
E\left[\ln \left( \frac{f_2(T)}{f_2(t)} \right) \right]\biggr|_{t = 0} &= \left(r - \eta_{2, 0}^2 \frac{s^2}{2} \right) T \\
Var\left[ \ln \left( \frac{f_2(T)}{f_2(t)} \right) \right]\biggr|_{t = 0} &= \eta_{2, 0}^2 s^2 T
\end{split}
\end{equation}

Whereby it follows that the same formula used to evaluate the ROV of $f_1(T_1)$ and $f_2(T_2)$ can be used to evaluate the ROV of $f_3(T_3)$. That is to say, Eq \eqref{eq:genEq} may be evaluated at $q = f_2(T_2)$ and $C = K_2$ to calculate $f_3$, the value of the option to invest in the second to last stage of research.

\begin{equation}
\begin{split}
f_3(f_2, t; T_2, K_2, s, r)|_{t = 0} &= e^{-r T_2} E[\max(f(T_2) - K_2, 0)]\bigr|_{t = 0} \\
&= f_2(0) \Phi \left(\delta_2 + \eta_{2,0} s \sqrt{T_2} \right) - e^{-r T_2} K_2 \Phi(\delta_2)
\end{split}
\end{equation}
<!-- \label{eq:rov3stage} -->

Where $T_2$ is the time until completion of stage 1, $K_2$ is the research investment required upon successful completion of stage 1---where stage 1 is considered successfully completed if $f_2(T_2) > K_2$---and where

\begin{equation}
\delta_2 = \frac{ \ln \left(\frac{f_2(0)}{K_2} \right) - \left(r - \eta_{2,0}^2 \frac{s^2}{2} \right) T_2} {\eta_{2,0} s \sqrt{T_2}}
\end{equation}

And so on and so forth, it follows generally that the ROV of stage 1 of any $n$-stage project, $f_n$, is an Itô process,

\begin{equation}
\Delta f_n = \left( \frac{\partial f_n}{\partial f_{n - 1}} r f_{n - 1} + \frac{\partial f_n}{\partial t} + \eta_{n - 1, 0}^2 \frac{s^2 f_{n - 1}^2}{2} \frac{\partial^2 f_n}{\partial f_{n - 1}^2} \right) \Delta t + f_n \eta_{n, 0} s \epsilon \sqrt{\Delta t}
\label{eq:itoLemN}
\end{equation}

And is thus lognormally distributed with parameters

\begin{equation}
\begin{split}
E\left[\ln \left(\frac{f_n(T)}{f_n(t)} \right) \right]\biggr|_{t = 0} &= \left(r - \eta_{n, 0}^2 \frac{s^2}{2} \right) T \\
Var\left[ \ln \left( \frac{f_n(T)}{f_n(t)} \right) \right]\biggr|_{t = 0} &= \eta_{n,0}^2 s^2 T
\end{split}
\:\:;\:\:\:  0 < T < T_{n - 1}
\label{eq:EnVarlFn}
\end{equation}

Such that a general formula for the ROV of stage 1 of any $n$-stage project may be resolved by evaluating Eq \eqref{eq:genEq} at $q = f_{n - 1}(T_{n - 1})$, $C = K_{n - 1}$, giving

\begin{equation}
\begin{split}
f_n(f_{n - 1}, t; T_{n - 1}, K_{n - 1}, s, r)|_{t = 0} &= e^{-r T_{n - 1}} E[\max(f_{n - 1}(T_{n - 1}) - K_{n - 1}, 0)]\bigr|_{t = 0} \:\:;\:\:\: n \in \mathbb{N}_0 \\
&= f_{n - 1}(0) \Phi \left(\delta_{n - 1} + \eta_{n - 1, 0} s \sqrt{T_{n - 1}} \right) - e^{-r T_{n - 1}} K_{n - 1} \Phi(\delta_{n - 1})
\end{split}
\label{eq:ROVn}
\end{equation}

Where $T_{n - 1}$ is the time until completion of stage 1 of the $n$-stage research project, $K_{n - 1}$ is the research investment required upon successful completion of stage 1 at time $T_{n - 1}$, where stage 1 is considered successfully completed if $f_{n - 1}(T_{n - 1}) > K_{n - 1}$, and where
<!-- ($T_{n - 1} < T_{n - 2} <...< T_0$) -->

\begin{equation}
\delta_{n - 1} = \frac{ \ln \left(\frac{f_{n - 1}(0)}{K_{n - 1}} \right) - \left(r - \eta_{n - 1, 0}^2 \frac{s^2}{2} \right) T_{n - 1}} {\eta_{n - 1, 0} s \sqrt{T_{n - 1}}}
\end{equation}

The ROV of stage 1 of any $n$-stage project is thus calculated recursively, starting with the calculation of the ROV of the last stage $f_1(x, t; T_0, K_0, s, r)|_{t = 0}$, which then feeds into the calculation of the ROV of the second to last stage, $f_2(f_1, t; T_1, K_1, s, r)|_{t = 0}$, and so forth until the calculation of first stage ROV $f_n(f_{n - 1}, t; T_{n - 1}, K_{n - 1}, s, r)|_{t = 0}$. The ROV $f_i$ calculated at each stage is the value of the option to invest in the subsequent stage.

\begin{equation}
\begin{split}
f_i(f_{i - 1}, t; T_{i - 1}, K_{i - 1}, s, r)|_{t = 0} &= e^{-r T_{i - 1}} E[\max(f_{i - 1}(T_{i - 1}) - K_{i - 1}, 0)]\bigr|_{t = 0} \\
&= f_{i - 1}(0) \Phi \left(\delta_{i - 1} + \eta_{i - 1, 0} s \sqrt{T_{i - 1}} \right) - e^{-r T_{i - 1}} K_{i - 1} \Phi(\delta_{i - 1})
\end{split}
\label{eq:ROVi}
\end{equation}

\begin{equation}
\delta_{i - 1} = \frac{ \ln \left(\frac{f_{i - 1}(0)}{K_{i - 1}} \right) - \left(r - \eta_{i - 1, 0}^2 \frac{s^2}{2} \right) T_{i - 1}} {\eta_{i - 1, 0} s \sqrt{T_{i - 1}}}
\end{equation}

In this process, the respective probabilities of finishing in the money (project stage success), $\Phi(\delta_i)$, and elasticities of $f_i$ with respect to project NPV, $\eta_{i, 0}$, are also calculated for each stage. The elasticity $\eta_{i, 0}$ is an instructive measure of ROV sensitivity to changes in project NPV. It is calculated for each project stage $i \in [0, n]$ as follows.

\begin{equation}
\eta_{i, 0} = \frac{f_{i - 1}}{f_{i}} \frac{\partial f_{i}}{\partial f_{i - 1}} \eta_{i - 1, 0} \:\:;\:\:\: 0 \leq i \leq n \:\:\: \eta_{0,0} = 1
\end{equation}

Where the derivative $\partial f_i / \partial f_{i - 1}$ follows from Eq \eqref{eq:ROVi}.

\begin{equation}
\frac{\partial f_i}{\partial f_{i - 1}} = \Phi \left( \delta_{i - 1} + \eta_{i - 1, 0} s \sqrt{T_{i - 1}} \right) + s \sqrt{T_{i - 1}} \frac{\partial \eta_{i - 1, 0}}{ \partial f_{i - 1}} f_{i - 1} \phi \left(\delta_{i - 1} + \eta_{i - 1, 0} s \sqrt{T_{i - 1}} \right)
\end{equation}

Where $\phi()$ is the standard normal probability density function, and

\begin{equation}
\frac{\partial \eta_{i - 1, 0}}{ \partial f_{i - 1}} =
\begin{cases}
\frac{\partial}{\partial f_{i - 1}} 1 = 0 &, \:\:\: i = 1 \\
\frac{\partial}{\partial f_{i - 1}} \left(\frac{f_{i - 2}}{f_{i - 1}} \frac{\partial f_{i - 1}}{\partial f_{i - 2}} \right) = -\frac{\eta_{i - 1}}{f_{i - 1}} &, \:\:\: i > 1
\end{cases}
\end{equation}

Finally, note that the multistage ROV formula may also be obtained in a "Black-Scholes manner" by recognizing that $f_n$ is defined as an expected future value discounted back to the present, such that the expected growth rate of $f_n$ is, by definition, just $E[\Delta f_n / f_n] = r \Delta t$. Since $f_n$ is an Itô process, then the expected growth rate is also given by

\begin{equation}
E\left[ \frac{\Delta f_n}{f_n} \right] = \frac{1}{f_n} \left( \frac{\partial f_n}{\partial f_{n - 1}} r f_{n - 1} + \frac{\partial f_n}{\partial t} + \eta_{n - 1, 0}^2 \frac{s^2 f_{n - 1}^2}{2} \frac{\partial^2 f_n}{\partial f_{n - 1}^2} \right) \Delta t
\end{equation}

Setting this equal to $r \Delta t$ results in a generalized Black-Scholes PDE for $f_n$,

\begin{equation}
rf_n = f_{n - 1} r \frac{\partial f_n}{\partial f_{n - 1}} + \frac{\partial f_n}{\partial t} + \eta_{n - 1, 0}^2 \frac{s^2 f_{n - 1}^2}{2} \frac{\partial^2 f_n}{\partial f_{n - 1}^2}
\end{equation}

Which may then be solved for $f_n$ under the boundary condition $f_n(T_n) = \max(f_{n - 1}(T_{n - 1}) - K_{n - 1}, 0)$ as another way of obtaining Eq \eqref{eq:ROVn}.

# An illustrative example

As an illustrative example, the multistage ROV formula developed above is applied to calculate the ROV a real multistage AR4D project currently in implementation. The aim of the project in question is to develop a transgenic potato variety with resistance to Late Blight disease (henceforth LBr potato), for release as a public good in two developing countries. Late Blight is the disease behind the infamous Irish Potato Famine of 1845-1849; and continues to pose a major threat to food security, especially in the developing world [@haverkort2008societal; @fry2008phytophthora]. Successful research and development of LBr potato varieties adapted to local environments is thus a potentially very high reward, disruptive proposition. It is also a high risk proposition, facing numerous research and non-research related challenges, not least of which is a strong "anti-GM" lobby in many of the target populations.

The LBr potato project is structured in four consecutive stages. These stages and their associated costs and time durations are summarized in Table 1, based on figures documented by Schiek et al. [-@schiek2016demys]. This is a purely pedagogical exercise designed to illustrate the usefulness of the multistage ROV model derived in the previous sections. While the exercise is based on real project stage costs and time horizons, the other parameter values are hypothetical. The resulting calculations should therefore not be interpreted as an authoritative valuation of LBr potato research. 
<!-- structure described in section \ref{sec:resStages}. The costs and time duration associated with each stage are summarized in Table 1,  -->
<!-- Costs are assessed as the sum of staff costs, direct operating costs (including lab bench costs), indirect operating costs (overhead), external contract costs, and stewardship costs. -->

```{r, echo = FALSE}
# Define functions
#----------------------------------------------------------------------------
# Table formatting function
# FitFlextableToPage <- function(ft, pgwidth = 6){
#   
#   ft_out <- ft %>% autofit()
#   #these_colWidths <- dim(ft_out)$widths * pgwidth  / (flextable_dim(ft_out)$widths)
#   these_colWidths <- dim(ft_out)$widths
#   these_colWidths[5] <- these_colWidths[1] * 1.5
#   #these_colWidths[6] <- these_colWidths[6] * 0.7
#   these_colWidths <- these_colWidths * pgwidth  / (flextable_dim(ft_out)$widths)
#   ft_out <- width(ft_out, width = these_colWidths)
#   return(ft_out)
# }

#============================================================================
#============================================================================
# End function definition
#============================================================================
#============================================================================
# Project stage time durations and costs (PREVIOUS DRAFT)
# (Values given in chronologically reverse order, starting with stage n.)
# Kvec[1] is the launch cost, Tvec[1] is the launch duration.
# Time durations given in years, multiplied by 4 to convert to quarters.
# MSU/cornell
# Tvec <- c(2, 3.5, 1, 2, 4) * 4
# Kvec <- c(300000, 200000, 400000, 400000, 530000)
# Tvec <- c(2, 3, 0.85, 1.7, 3.4) * 4
# Kvec <- c(250000, 181000, 312000, 336000, 530000)
# Kvec <- c(180541, 311974, 335530, 530250)
# CIP
# Kvec <- c(52000, 213000, 396000, 929000)
# Tvec <- c(1, 0.25, 2, 4.75) * 4
#===============================================================================
# Stage time duration params (CURRENT DRAFT)
# Stage durations - last stage first
# periodDurations <- c(3, 4, 2, 3) * 4
# tauVec <- rev(cumsum(periodDurations))
#TvecYrs <- c(4, 2, 4, 4); TvecQtrs <- TvecYrs * 4 # last first
#TvecYrs <- c(3, 1, 2, 4) # last first
TvecYrs <- c(4, 2, 1, 3); TvecQrs <- TvecYrs * 4 # In order
tauVecYrs <- rev(cumsum(TvecYrs)); tauVecQrs <- tauVecYrs * 4 # reverse so last first
stage1Invest <- 531000; launchCost <- 500000
Kvec <- c(launchCost, 181000, 312000, 336000, stage1Invest) # Last stage first
#===============================================================================
# Define discount rate
rYrly_discrete <- 0.12 #0.035
rQrly_discrete <- (1 + rYrly_discrete)^(1 / 4) - 1
rYrly <- round(log(1 + rYrly_discrete), 3)
r <- round(log(1 + rQrly_discrete), 3)
#===============================================================================
# Create table summarizing stage costs, time durations, and descriptions
stage1desc <- "Basic replication and scaling up"
stage2desc <- "Multi-location/season testing"
stage3desc <- "Compilation of the regulatory dossier"
stage4desc <- "Deregulation"
stageLaunchDesc <- "Distribution/outscaling/marketing of finished research product"
#launchDesc <- "Launch in 2 countries"
descVec <- c(stage1desc, stage2desc, stage3desc, stage4desc, stageLaunchDesc)

df_table <- data.frame(Stage = c(1:4, "Launch"),
                       Kn = rev(Kvec),
                       Duration = c(TvecQrs, "--"),
                       DurationCum = c(rev(tauVecQrs), "--"),
                       Description = descVec)

colnames(df_table)[2:4] <- c("Cost\n(USD)", "Duration", "Time until\ncompletion")

#df_table <- regulartable(df_table)
#df_table
#df_table <- width(df_table, width = 0.7)
table_title <- "Project to research and develop Late Blight resistant potato for release as a public good in two developing countries."
table_caption <- "Time expressed in annual quarters. Stage 1-4 costs and durations based on figures reported by Schiek et al. (2016)."# The launch cost and duration is hypothetical."

#df_table <- FitFlextableToPage(df_table)
# df_table <- add_header_lines(df_table, table_title)
# df_table <- add_footer_lines(df_table, table_caption)
# modifiedColnames <- colnames(df_table)
# modifiedColnames[3] <- "Duration\n(annual quarters)"
df_table %>%
  knitr::kable(
    format = "latex",
    caption = table_title,
    align = "l",
    booktabs = T,
    #longtable = T,
    linesep = " ",
    escape = F
  ) %>%
  kableExtra::kable_styling(
    position = "left",
    latex_options = c("HOLD_position", "scale_down", "striped", "repeat"),
    full_width = F,
    stripe_color = "gray!15"
  ) %>%
  # kableExtra::column_spec(column = 3:4, width = "1.5in"
  # ) %>%
  kableExtra::footnote(
    general = table_caption)
# number = c("Footnote 1; ", "Footnote 2; "),
# alphabet = c("Footnote A; ", "Footnote B; "),
# symbol = c("Footnote Symbol 1; ", "Footnote Symbol 2"))

# print_this_table <- function(df_table){
# df_table <- regulartable(df_table)
# df_table <- width(df_table, width = 5.5)
# # df_table <- FitFlextableToPage(df_table)
# 
# df_table %>% #regulartable() %>% autofit() %>%
#   valign(valign = "top", part = "all") %>%
#   align(align = "left", part = "all") %>%
#   fontsize(i = NULL, j = NULL, size = 9, part = "body") %>%
#   fontsize(i = NULL, j = NULL, size = 10, part = "header")
#   
# }


```

The donor, USAID, originally awarded the LBr potato research contract to Cornell University in 2010, with planned release in India. However, stage 1 research and development at Cornell was unsuccessful. In 2015, USAID transferred the contract to Michigan State University (MSU), which proposed a different stage 1 research strategy, with planned release in Bangladesh and Indonesia instead of India. India was dropped as a target country partly because of the vigorous anti-GM lobby there, which made stage 4 success unlikely. MSU completed stage 1 successfully in 2019, and a couple years later USAID awarded them a new contract for stage 2 research and development.

In the multistage real options language developed above, USAID bought an option on stage 2 of a 4 stage LBr potato research project when it awarded the original contract to Cornell; but this option expired out of the money at the end of stage 1. USAID then again bought an option on stage 2 research when it transferred the contract to MSU. This time, the option expired in the money, triggering investment in stage 2 research. The investment in stage 2 research is an option on stage 3.
<!-- , and is now approaching expiry in the money. -->

The aim of the present exercise is to calculate the ROV of stage 1 of the 4-stage LBr potato research project that USAID awarded to MSU in 2015. This is the value of USAID's option to fund stage 2 once stage 1 is complete. The exercise is conducted from the donor's perspective in the months prior to the "go/no-go" stage 1 investment decision. All values are given in terms of 2015 US dollars.

In order to make the calculation, the research stage and final launch costs $K_0, K_1, \dots, K_4$ and time horizons $T_0, T_1, \dots, T_3$ listed in Table 1 are required, as well as the quarterly discount rate $r$, the project NPV volatility $s$, and the project NPV at the time of the go/no-go decision to invest in stage 1 LBr research, $x(0)$.

USAID generally uses a high annual discount rate of $0.12$ in its cost-benefit analyses of agricultural development projects [@usaid2015cba]. A high discount rate is reasonable considering the long time horizon and high uncertainty surrounding project NPV, and considering that any alternative use of USAID funds would entail a similar level of elevated risk. The same annual (discrete) discount rate is thus adopted in the present exercise, which works out to a (continuous) quarterly discount rate of `r r` used in the calculation.

```{r, fig.show = "hold", fig.width = 5, fig.height=4, fig.align="center", fig.cap="\\label{fig:netBens}(Top) A hypothetical 12 year projection of the discounted and undiscounted LBr potato net benefit stream from time of release in the target countries. (Bottom) The integral of discounted net benefits up to each year.", echo = FALSE}
# Define functions
get_di <- function(K_im1, f_im1, tau_im1, m, s, eta_im10 = 1){
  d_i <- (log(f_im1 / K_im1) + (m - eta_im10^2 * s^2 / 2) * tau_im1) / (s * eta_im10 * sqrt(tau_im1))
  return(d_i)
}
ROVi <- function(K_im1, f_im1, tau_im1, m, s, eta_im10 = 1){
  d_i <- get_di(K_im1, f_im1, tau_im1, m, s, eta_im10)
  u <- d_i + eta_im10 * s * sqrt(tau_im1)
  rov <- f_im1 * pnorm(u) - exp(-r * tau_im1) * K_im1 * pnorm(d_i)
  probSuc <- pnorm(d_i)
  return(c(rov, probSuc))
}
get_dfidfim1 <- function(K_im1, f_im1, f_i, tau_im1, m, s, eta_im10 = 1){
  d_i <- get_di(K_im1, f_im1, tau_im1, m, s, eta_im10)
  u <- d_i + eta_im10 * s * sqrt(tau_im1)
  if(eta_im10 == 1){dEtaim10dfim1 <- 0}else{dEtaim10dfim1 <- -eta_im10 / f_im1}
  dfidfim1 <- pnorm(u) + dEtaim10dfim1 * s * sqrt(tau_im1) * f_im1 * dnorm(u)
  if(abs(dfidfim1) < 10^-10){dfidfim1 <- 0}
  return(dfidfim1)
}
ROVn <- function(X0, Kvec, tauVec, m, s){
  f_im1 <- X0; eta_im10 <- 1
  n <- length(Kvec)
  fiOut <- c(); etaOut <- c(); probSucVec <- c()
  for(i in 1:n){
    tau_im1 <- tauVec[i]; K_im1 <- Kvec[i]
    outROVi <- ROVi(K_im1, f_im1, tau_im1, m, s, eta_im10)
    f_i  <- outROVi[1]; probSuc <- outROVi[2]
    dfidfim1 <- get_dfidfim1(K_im1, f_im1, f_i, tau_im1, m, s, eta_im10)
    eta_i0 <- f_im1 / f_i * dfidfim1 * eta_im10
    fiOut[i] <- f_i; etaOut[i] <- eta_i0; probSucVec[i] <- probSuc
    f_im1 <- f_i; eta_im10 <- eta_i0
  }
  return(data.frame(f_i = fiOut, eta_i0 = etaOut, probSuc = probSucVec))
}
rootROVn <- function(X0, Kvec, tauVec, mm, s, stage1Invest){
  dfOut <- ROVn(X0, Kvec, tauVec, mm, s)
  f_n <- dfOut$f_i[nrow(dfOut)]
  slack <- f_n - stage1Invest
  return(slack)
}
#===============================================================================
#2013 potato VoP 2014-2016 constant USD
VoPIndonesia <- 0.8 #Indonesia: 0.8 billion
VoPBangladesh <- 1.6 #Bangladesh 1.6 billion
VoPKenya <- 0.8 #Kenya: 0.8 billion
VoPNigeria <- 0.5 #Nigeria: 0.5 billion
VoPvec <- c(VoPIndonesia, VoPBangladesh, VoPKenya, VoPNigeria)
# VoPtot <- sum(VoPvec)
# VoPpot <- VoPtot / 0.9
# yrlyLoss <- VoPpot - VoPtot
thisVoP <- VoPNigeria
VoPpot <- thisVoP / 0.9
yrlyLoss <- VoPpot - thisVoP
#yrlyLoss
# yrlyLoss <- VoPpot * 0.15 #billion
#yrlyLoss <- VoPIndonesia / 0.9 *  #billion
#===============================================================================
# Calculate project NPV(Tn) i.e. X0
yrHoriz <- 12; qtrHoriz <- yrHoriz * 4
t <- seq(0, qtrHoriz, length.out = qtrHoriz + 1)
# alphaYrly_discrete <- 0.65
# alphaQrly_discrete <- (1 + alphaYrly_discrete)^(1 / 4) - 1
# alphaQrly <- round(log(1 + alphaQrly_discrete), 3)
alphaQrly <- 0.6
dNVT0ref <- 0.11; tRef <- 12
dNVT0 <- dNVT0ref * (t / tRef)^alphaQrly
dNPVT0_discrete <- dNVT0 / (1 + rQrly_discrete)^t
#dNPVT0_discrete <- dNVT0 / (1 + rYrly_discrete)^t
dNPVT0_contin <- exp(-r * t) * dNVT0
#dNPVT0_contin <- exp(-rYrly * t) * dNVT0
library(expint)
A <- dNVT0ref * tRef^(-alphaQrly)
NPVT0 <- A / r^(1 + alphaQrly) * (-gammainc(1 + alphaQrly, r * t) + gammainc(1 + alphaQrly, 0))
NPVT0_discrete <- cumsum(dNPVT0_discrete)
NPV0_discrete <- exp(-r * tauVecQrs[1]) * NPVT0_discrete
NPV0 <- exp(-r * tauVecQrs[1]) * NPVT0
#NPV0 <- exp(-rYrly * tauVecYrs[1]) * NPVT0
# plot(t, dNVT0)
# plot(t, dNPVT0)
# plot(t, NPVTn)
#-------------------------------------------------------------------------------
nbCutoffYr <- 12 # Net benefit cutoff year
nbCutoffQtr <- nbCutoffYr * 4
X0 <- NPV0[nbCutoffQtr] * 10^6
#-------------------------------------------------------------------------------
# Plot dNV, dNPV, NPV
library(patchwork)
# Set graphic parameters
axisTitleSize <- 7
axisTextSize <- 6
legendKeySize <- 0.3
legendTextSize <- axisTextSize
#facetTitleSize <- axisTextSize
#plotTitleSize = axisTextSize
#labelSize <- 2
# Organize plot data into data frames
dfdNVT0 <- data.frame(Quarter = t, val = dNVT0, type = "Undiscounted\nnet benefits")
dfdNPVT0 <- data.frame(Quarter = t, val = dNPVT0_contin, type = "Continuously discounted\nnet benefits")
dfNPVT0 <- data.frame(Quarter = t, NPV = NPVT0)
df1 <- data.frame(rbind(dfdNVT0, dfdNPVT0))
df2 <- dfNPVT0
theseColors <- randomcoloR::distinctColorPalette(2)
# Plot undiscounted and discounted net benefits overlaid
gg <- ggplot(df1, aes(x = Quarter, y = val, fill = type))
gg <- gg + geom_bar(stat = "identity", position = position_dodge(width = 0.25), color = "black")
gg <- gg + scale_fill_manual(values = theseColors)
gg <- gg + scale_x_continuous(breaks = seq(0, qtrHoriz, 4),
                              labels = seq(0, yrHoriz, 1))
gg <- gg + labs(x = "Years from release", y = "Million USD")
gg <- gg + theme_bw()
gg <- gg + theme(axis.text = element_text(size = axisTextSize),
                 axis.title = element_text(size = axisTitleSize),
                 legend.title = element_blank(),
                 legend.position = "top",
                 legend.text = element_text(size = legendTextSize),
                 legend.key.size = unit(legendKeySize, "cm"))
gg1 <- gg
# Plot NPV
gg <- ggplot(dfNPVT0, aes(x = Quarter, y = NPV))
gg <- gg + geom_bar(stat = "identity", color = "black")
gg <- gg + scale_x_continuous(breaks = seq(0, qtrHoriz, 4),
                              labels = seq(0, yrHoriz, 1))
gg <- gg + labs(x = "Years from release", y = "NPV\n(million USD)")#, title = "(c)")
gg <- gg + theme_bw()
gg <- gg + theme(axis.text = element_text(size = axisTextSize),
                 axis.title = element_text(size = axisTitleSize))#,
#                 plot.title = element_text(size = axisTitleSize))
gg2 <- gg
gg1 / gg2 #+ plot_annotation(tag_levels = "a")
#===============================================================================
# Define project NPV volatility:
cv <- 1.5
s <- cv * r * sqrt(tauVecQrs[1])
# Calculate ROV of stage 1 research
m <- r
dfROVn <- ROVn(X0, Kvec[-length(Kvec)], tauVecQrs, m, s)
f_n <- dfROVn$f_i[nrow(dfROVn)]
#f_n - stage1Invest
# Conventional CBA
CBAlhs <- X0 - sum(exp(-r * tauVecQrs) * Kvec[-length(Kvec)])
CBA <- CBAlhs - stage1Invest
#===============================================================================
# "Manual" check step by step
eta_00 <- 1; f0 <- X0
outROVi <- ROVi(Kvec[1], f0, tauVecQrs[1], m, s, eta_00) # Value of option to launch
f1 <- outROVi[1]; probSuc1 <- outROVi[2]
df1df0 <- get_dfidfim1(Kvec[1], f0, f1, tauVecQrs[1], m, s, eta_00)
#pnorm(get_di(Kvec[1], X0, tauVecQrs[1], m, s, eta_00) + eta_00 * s * sqrt(tauVecQrs[1]))
eta_10 <- f0 / f1 * df1df0 * eta_00
outROVi <- ROVi(Kvec[2], f1, tauVecQrs[2], m, s, eta_10) # Value of option on stage 4 research
f2 <- outROVi[1]; probSuc2 <- outROVi[2]
df2df1 <- get_dfidfim1(Kvec[2], f1, f2, tauVecQrs[2], m, s, eta_10)
eta_20 <- f1 / f2 * df2df1 * eta_10
outROVi <- ROVi(Kvec[3], f2, tauVecQrs[3], m, s, eta_20) # Value of option on stage 3 research
f3 <- outROVi[1]; probSuc3 <- outROVi[2]
df3df2 <- get_dfidfim1(Kvec[3], f2, f3, tauVecQrs[3], m, s, eta_20)
eta_30 <- f2 / f3 * df3df2 * eta_20
outROVi <- ROVi(Kvec[4], f3, tauVecQrs[4], m, s, eta_30) # Value of option on stage 2 research
f4 <- outROVi[1]; probSuc4 <- outROVi[2]
df4df3 <- get_dfidfim1(Kvec[4], f3, f4, tauVecQrs[4], m, s, eta_30)
eta_40 <- f3 / f4 * df4df3 * eta_30
#-------------------------------------------------------------------------------
dfROVnCheck <- data.frame(f_i = c(f1, f2, f3, f4),
                          eta_i0 = c(eta_10, eta_20, eta_30, eta_40),
                          probSuc = c(probSuc1, probSuc2, probSuc3, probSuc4))
#dfROVn - dfROVnCheck


```
<!-- Ex-ante impact assessments of LBr potato in the target countries are underway, but have not yet been published at the time of writing [pers correspondence]. -->

At the 2015 go/no-go decision point, no published ex-ante impact assessments of LBr potato in the target countries exist. For the pedagogical purposes of this exercise, I suppose that the donor conducts its own ex-ante impact assessment internally, finding that annual net benefits are likely to scale sublinearly with time by `r alphaQrly`% for every $1$% increase in time, reaching \$`r dNVT0ref * 10^3` thousand in year `r tRef / 4` after release. Mathematically, this may be expressed

\begin{equation}
y(t) = y(\hat{t}) \left( \frac{t}{\hat{t}} \right)^{\alpha}
\end{equation}

Where $y(t)$ stands for LBr potato net benefit at $t$ annual quarters after release, $y(\hat{t}) =$ \$`r format(dNVT0ref * 10^6, scientific = F)`, $\alpha =$ `r alphaQrly`; and $\hat{t} =$ `r tRef` quarters (i.e., 3 years).

The corresponding continuously discounted net benefit stream ($e^{-rt} y(t)$) is overlaid onto the undiscounted stream ($y(t)$) in the top panel of Fig \ref{fig:netBens}. (Again, this is a hypothetical projection for pedagogical purposes. A recent ex ante impact assessment suggests that net benefits could be substantially higher [@kihiu2025ex].) Integration of the discounted net benefits with respect to $t$ over $[T_0, T_{-1}]$ gives project NPV at time of release ($x(T_0)$).

\begin{equation}
\begin{split}
x(T_0) &= \int_{T_0}^{T_{-1}} e^{-rt} y(t) \: dt \\
&= \frac{y(\hat{t})}{\hat{t}^{\alpha} r^{\alpha + 1}} ( -\Gamma(\alpha + 1, r(T_{-1} - T_0)) + \Gamma(\alpha + 1, 0) )
\end{split}
\end{equation}
<!-- x(T_0)|_{t = T_{-1}} -->

Where $\Gamma(u,v)$ is the incomplete gamma function. The upper bound $T_{-1}$ is a net benefit cutoff set by the donor. Net benefits continue beyond $T_{-1}$, but the donor is only interested in net benefits up to $T_{-1}$. Values of $x(T_0)$ are given in the bottom panel of Fig \ref{fig:netBens} for cutoffs ranging from $1$ to `r nbCutoffYr` years after release. For the calculation of ROV in the present hypothetical exercise, the donor sets $T_{-1}$ to `r nbCutoffYr` years (`r 4 * nbCutoffYr` quarters) after release. This gives a project NPV of \$`r round(NPVT0[nbCutoffQtr], 3)` million at time of release $T_0$, i.e. at the end of stage 4 research. The project NPV discounted back to the time of the donor's go/no-go decision to fund stage 1 research is then $x(0) = e^{-rT_{4}} x(T_0) =$ \$`r round(X0 * 10^-6, 3)` million.

The only remaining information required to calculate ROV is the LBr potato release and outscaling cost due at $t = T_0$ upon successful completion of stage 4, and the project NPV volatility parameter $s$. For this pedagogical exercise, it is hypothesized that the donor estimates the release and outscaling cost to be \$`r Kvec[1] * 10^-3` thousand. Because no project NPV time series exists whereby project volatility $s$ might be directly estimated, an implicit value for $s$ must be deduced. One practical way to to do this is calculate $s$ based on expert elicitation of the upper and lower bounds on $x(T_0)$. That is to say, interpreting these bounds as the 95% confidence interval, a value for $s$ can be "backed out" from the assumption that project NPV follows a gBm (which implies lognormally distributed $x(T_0)$). In the present exercise, I hypothesize that such a process results in a value of $s=$ `r round(s, 3)`. This corresponds to a coefficient of variation ($s / (r \sqrt{T_0})$) of `r cv`, reflecting the high uncertainty that typically surrounds the expected impact of disruptive new technology a decade in advance of its completion and release.
<!-- \ref{sec:gBmDefens} -->

With all of the necessary information in hand, the ROV of stage 1 LBr research is calculated by first calculating the ROV of stage 4, the last stage, and then calculating the ROV of stage 3, and so forth, recursively, as explained in the previous section. In the present exercise, this results in a stage 1 LBr research ROV of \$`r format(round(f_n), scientific = F)`, which exceeds the stage 1 cost of \$`r format(stage1Invest, scientific = F)` by \$`r format(round(f_n - stage1Invest), scientific = F)`, resulting in a "go" investment decision. On the ROV basis, then, the stage 1 investment is justified. By the conventional CBA criterion, meanwhile, the project would be rejected, as the left-hand side of inequality \eqref{eq:cbaCrit} works out to \$`r format(round(CBAlhs), scientific = F)`, which falls below the stage 1 cost by \$`r format(abs(round(CBA)), scientific = F)`.
<!-- \ref{sec:method} -->
<!-- \begin{equation} -->
<!-- 1180000 - \sum_{i = 0}^3 e^{-r T_i} K_i < 530000 \:\: \rightarrow \:\: \text{reject project} -->
<!-- \label{eq:convCBA} -->
<!-- \end{equation} -->

The recursively calculated ROVs for each stage of research are presented in Table 2. Also included in this table are each stage's ROV elasticity with respect to project NPV and probability of finishing in the money, which are calculated in the process. This provides additional information, not provided in conventional CBA, that the donor may find useful in their decision making process.

```{r, echo=FALSE}
#===============================================================================

dfROVn$Stage <- 4:1; #dfROVn$i <- 1:4
dfROVn <- dfROVn[, c("Stage", "f_i", "eta_i0", "probSuc")]
dfROVn$f_i <- round(dfROVn$f_i)
dfROVn$eta_i0 <- round(dfROVn$eta_i0, 3)
dfROVn$probSuc <- round(dfROVn$probSuc, 3)
dfROVn <- dfROVn[order(dfROVn$Stage, decreasing = F), ]
table_title <- "LBr potato project stage 1-4 ROV"
colnames(dfROVn) <- c("Stage", "ROV", "ROV elasticity w.r.t. project NPV", "Probability of success")
dfROVn %>%
  knitr::kable(
    format = "latex",
    row.names = F,
    caption = table_title,
    align = "l",
    booktabs = T,
    #longtable = T,
    linesep = " ",
    escape = F
  ) %>%
  column_spec(2:4, width = "4cm") %>%
  kable_styling(
    position = "center",
    latex_options = c("HOLD_position", "striped", "repeat"),
    full_width = F,
    stripe_color = "gray!15",
    font_size = 8
  )
#%>%
#  kableExtra::column_spec(column = 4:7, width = "1in")



# df_table %>%
#   knitr::kable(
#     format = "latex",
#     caption = table_title,
#     align = "l",
#     booktabs = T,
#     #longtable = T,
#     linesep = " ",
#     escape = F
#   ) %>%
#   kableExtra::kable_styling(
#     position = "left",
#     latex_options = c("HOLD_position", "scale_down", "striped", "repeat"),
#     full_width = F,
#     stripe_color = "gray!15"
#   ) %>%
#   # kableExtra::column_spec(column = 3:4, width = "1.5in"
#   # ) %>%
#   kableExtra::footnote(
#     general = table_caption)


#===============================================================================
# Find breakeven project NPV
#rootROVn(X0 * 0.2, Kvec, tauVecQrs, mm = r, s, stage1Invest)
thisInt <- c(10^5, 10^8)
outUniroot <- uniroot(rootROVn, interval = thisInt,
                                 #lower = min(thisInt), upper = max(thisInt),
                                 Kvec = Kvec[-length(Kvec)], tauVec = tauVecQrs,
                                 mm = r, s = s, stage1Invest)
X0min <- outUniroot$root
#ROVn(X0min, Kvec, tauVecQrs, m, s)

```

In cases where no ex-ante impact assessment of project NPV is available, it may be instructive to solve ROV of stage 2 research for the break even project NPV, i.e. the value of $x(0)$ required to yield an ROV equal to the stage 1 cost. In this example, the break even project NPV is \$`r round(X0min * 10^-6, 3)` million. Any project NPV above this value results in an ROV of stage 2 LBr research greater than the investment required to implement stage 1, and thus in a go decision.

# Discussion and conclusion

The ROV method presented in this article is novel in two regards. First of all, it accommodates the multistage structure of AR4D projects. The key step in the derivation permitting this development is the demonstration via Itô's lemma that if project NPV is lognormally distributed, then ROV itself is also lognormally distributed with computable parameters given in Eqs \eqref{eq:EnVarlFn}. Since the single stage ROV formula applies to any lognormally distributed variable, ROV itself may then be fed back into this formula to evaluate the ROV of ROV; and this may be repeated to evaluate the stage 1 ROV of a project of any number of stages. In the process of computing the stage 1 ROV of a multistage project, the ROV of all subsequent project stages, together with their respective probability of success and sensitivity to changes in the underlying project NPV, are also computed, providing the donor with potentially useful complementary information on which to base the strategic go/no-go funding decision for stage 1.

Secondly, whereas most ROV methods build off of the foundations of financial option pricing---often leading to strained attempts to construe projects as traded goods in efficient markets---here the method is built upon suitable non-market foundations. These non-market foundations are considerably simpler than their financial counterparts. Instead of the replicating portfolios/complete markets and no-arbitrage rule assumed in financial and corporate contexts---assumptions which themselves rest upon broader conditions regarding competitive market equilibrium and informational efficiency---the method presented here rests on two required conditions: 1) that NPV and ROV are expected to grow at the discount rate (something that is definitionally true and so does not need to be assumed); and 2) that future project NPV is approximately lognormally distributed through time---i.e., that the evolution of project NPV is reasonably well approximated by gBm.

To justify the latter condition, which is often contested in the ROV literature, I argue that the discontinuities practitioners seek to accommodate through jump or jump-diffusion models are often artifacts of reporting protocols rather than features of the value-generating process that is reported. In the AR4D context, the relevant object of analysis is not the discretely reported value trajectory, but the latent evolution of expected social value itself, which may plausibly be treated as continuous. I further address the related concern that project value may be path dependent, and hence not well approximated by a Markov process. While historical contingencies undoubtedly shape underlying technological and institutional dynamics, standard results in financial economics and valuation practice suggest that, under rational updating, current value may be treated as a sufficient summary of decision-relevant information for practical purposes. Together, these considerations clarify the conditions under which the latent project value process may be reasonably approximated as continuous and Markovian. Finally, even if substantial discontinuities or path-dependent effects arise over the full 10-20 year horizon of a typical AR4D project, the multistage framework substantially shortens the effective valuation horizon over which these approximations must hold, typically to sequential decision intervals on the order of 1-5 years.

The multistage ROV model derived in this article is not limited to transgenic projects of the type explored in the illustrative example above, but may be applied more broadly in any non-market research setting where the conditions just described are plausibly met. This includes conventional crop and livestock breeding projects, and also potentially extends to projects not necessarily focused on genetic gain, or that involve a considerable socio-ecological component---e.g., agroecology, climate-smart agriculture, value chain integration, and environmental services---so long as the stated conditions remain defensible.
<!-- The requirement that project NPV approximately follow a gBm is, more specifically, a requirement that proportional changes in the latent project value process be individually small, collectively numerous, and approximately independent. -->
<!-- The requirement that project NPV approximately follow a gBm is, more specifically, a requirement that changes in project NPV be continuous, individually small, collectively numerous, and approximately independent, and where current project fundamentals provide a reasonably sufficient summary of decision-relevant historical information. The prevailing view in much of the corporate ROV literature is that project NPV is discontinuous and path dependent. In the introductory defense of gBm I argue that apparent discontinuities are artifacts of reporting protocols, discuss the conditions under which the latent project value may be considered continuous; and why project value may be reasonably interpreted as effectively summarizing all decision relevant historical information. I also note that while substantial discontinuities may occasionally arise over the full 10–20 year horizon of a typical AR4D project due to extreme events (wars, market collapses, serendipitous discoveries, etc.), the proposed multistage framework substantially shortens the effective valuation horizon over which the continuous approximation must remain empirically plausible---typically to sequential decision intervals on the order of 1–5 years. -->
<!-- The multistage ROV model derived in this article is not limited to transgenic projects of the type explored in the illustrative example above, but may also be applied in the valuation of conventional crop and livestock breeding projects; and more generally in any research setting where changes in _latent_ project value follow a gBm---i.e., where a meaningful distinction may be drawn between a reported value series and the latent value series that is reported, and where the latter is plausibly continuous, individually small, collectively numerous, and approximately independent, and where current project fundamentals provide a reasonably sufficient summary of decision-relevant historical information. As discussed in the introductory defense of gBm, these conditions are met in a susprising range of projects, potentially including multistage projects not necessarily built around the objective of genetic gain and/or involving a considerable socio-ecological dimension (e.g., agroecology, climate smart agriculture, value chain integration, environmental services). -->
<!-- The multistage ROV model derived in this article is not limited to transgenic projects of the type explored in the illustrative example above, but may be applied more broadly in any AR4D research setting where the conditions just described are plausibly met. This requires, first, that a meaningful distinction can be drawn between a reported value series and the latent value process it periodically reflects — a distinction that becomes more rather than less plausible as the project encompasses a wider range of heterogeneous ecological, technical, political, economic, and social processes whose aggregate behavior tends toward continuity. It requires, second, that current project fundamentals provide a reasonably sufficient summary of decision-relevant historical information, so that forward-looking valuation does not depend materially on the full historical path. Where these conditions are met, the model may be applied to conventional crop and livestock breeding projects, and more generally to multistage projects involving a considerable socio-ecological dimension — including agroecology, climate-smart agriculture, value chain integration, and environmental services — so long as proportional changes in latent expected project value are individually small, collectively numerous, and approximately independent, and current project fundamentals provide a reasonably sufficient summary of decision-relevant historical information. -->
<!-- derivation at no point implies all the problematic assumptions of financial/corporate option value... -->
<!-- tradeoff--increasing volatility increases ROV but decreases probability of success... -->
<!-- abstract from the messiness... in particular deadline extensions, staggered finishing (Kenya) and release -->
<!-- Assessment of $s$... -->
<!-- potential use as insurance... -->
<!-- ...can be used in any other suitable non-ar4d context -->
<!-- The proposed model can also be applied at more aggregate levels of accounting---to valuate, for example, a multistage program consisting of several projects in various stages of implementation. -->
<!-- As for the relation between project NPV and the outscaling investment, I refer to the marginally diminishing relationship that typically exists between AR4D project NPV and the size of the investment in the outscaling of the finished technology. That is to say, beyond a certain minimum level of investment required to release and support the diffusion and uptake of the project's research product on a pilot level, further increments in the outscaling investment have the effect of broadening the scope of impact---i.e. the size and number of target populations and environments where the research product is released and has an impact. Existing ROV methods tacitly assume that project NPV and the outscaling investment are independent, or that only one outscaling investment scenario is of interest. The adjustment introduced here facilitates examination of all outscaling investment scenarios, and even makes it possible to solve for the ROV-maximizing outscaling investment. -->
<!-- In this article, I have developed an $n$-fold ROV model to evaluate the real option value of multistage AR4D projects. The model effectively prices in the research donor's option to discontinue funding of a project at the end of well defined research stages, thereby facilitating donor commitment to the long time horizons of agricultural research. In other words, the model formalizes the de facto structure of most AR4D project funding arrangements. The model also accounts for the wide uncertainty that usually surrounds AR4D project NPV, thereby reducing the pressure on researchers to put an exact monetary value on impacts that may be decades in the future. I have also extended the model to include abandonment value. -->
<!-- The proposed model is particularly appropriate for the valuation of high risk, deep-in-the-money (i.e. high expected reward) projects. The higher a project's moneyness (i.e. the higher the expected reward), the higher must be its volatility for there to be a meaningful difference between the ROV and conventional CBA approaches. In the case of LBr potato examined above, for example, a small increase in the hypothetical NPV and/or decrease in the confidence interval given by $\bar{\ell}$, $\underline{\ell}$ (and hence a decrease in the volatility $s$) results in an ROV that is virtually the same as the NPV, in which case there is no point in evaluating the $n$-fold ROV. -->
<!-- The proposed model rests on the assumption that project NPV follows a gBm, and is thus subject to criticism insofar as this assumption is unrealistic. In response to critics who reject the gBm assumption on the basis that the time evolution of project NPV fails to resemble that of gBm, I have argued that 1) the time domain discrepancy is largely spurious, attributable to one's theory of how/when new information affects project NPV, and to one's choice of time step $\Delta t$. And, more importantly, I have argued that 2) it is not the time domain, but the frequency domain that is relevant when it comes to far-from-market ROV calculation. That is to say, the calculation of ROV hinges upon the size distribution of changes in NPV, regardless of when or in what order the changes occur. Assuming that project NPV follows a gBm is tantamount to assuming that percentage changes in project NPV are normally distributed. Insofar as this assumption is valid, then, the proposed model is valid. Practitioners are forewarned that rejection of this assumption tends to lead to the sort of black box complexity that has so far inhibited wider adoption of the ROV approach. -->
<!-- Note that project NPV usually scales with the launch and scaling up cost $K_n$. That is to say, beyond a certain minimum level of funding required to generate, release, and support the diffusion and uptake of the project's research product on a pilot level, further increments in the investment $K_n$ have the effect of broadening the scope of impact---i.e. extending the size and number of target populations and environments where the research product is released and has an impact. In future work, then, it might be worthwhile to explore the ROV implications of explicitly modeling project NPV as a function of the scaling up investment $K_n$. For example, $x(0)$ could be defined -->
<!-- \begin{equation} -->
<!-- x(0) = \tilde{x}(0) \left(\frac{K_n}{\bar{K}_n}\right)^{\alpha} -->
<!-- \end{equation} -->
<!-- Where the parameter $\bar{K}_n$ is a reference value---perhaps the minimum or maximum possible scale up investment, for example---$\tilde{x}(0)$ is the project NPV if scaling up funding equals the reference value, and the exponent $\alpha$ is restricted to fall between 0 and 1, reflecting decreasing returns to scale. It may then be interesting to analyze ROV response to successively higher scale up investments $K_n$, corresponding to successively broader geographical scopes of impact. -->

# Supporting information {-}

**S1 Appendix.** A) Proof of Eq \eqref{eq:genEq} through straightforward integration, and B) Derivation of Eq \eqref{eq:rov} from the Black-Scholes PDE.

# References {-}

<div id="refs"></div>
<!-- ::: {#refs} -->
<!-- ::: -->

\newpage

<!-- \appendix -->
<!-- \counterwithin*{equation}{section} -->
<!-- \renewcommand\theequation{\thesection\arabic{equation}} -->
\setcounter{equation}{0}
\renewcommand{\theequation}{A\arabic{equation}}

# Appendix {-}
<!-- # Appendix {-} -->

<!-- \appto\appendix{\counterwithin{equation}{\Alph{section}}} -->
<!-- \appendix -->
<!-- \section{Proof of Eq \eqref{eq:genEq} through straightforward integration} -->
<!-- \renewcommand{\thesubsection}{\Alph{subsection}} -->
<!-- \renewcommand{\thesubsection}{A} -->

<!-- \numberwithin{equation}{section} -->
<!-- \makeatletter -->
<!-- % "activate" the preparatory code, but for section-level headers only -->
<!-- \newcommand{\section@cntformat}{Appendix \thesection:\ } -->
<!-- \makeatother -->

<!-- \numberwithin{equation}{section} -->
<!-- \renewcommand{equation}{\Alph{section}} -->

## Proof of Eq \eqref{eq:genEq} through straightforward integration {-}
<!-- Lemma: -->

<!-- If $v(t)$ follows a geometric Brownian motion; that is to say, if the instantaneous evolution of $v(t)$ over time can be expressed -->

<!-- \begin{equation} -->
<!-- \Delta v = \alpha v \Delta t + \beta v \epsilon \sqrt{\Delta t} \:\:;\:\:\: \Delta v = v(t + \Delta t) - v(t) -->
<!-- \label{eq:deltaV} -->
<!-- \end{equation} -->

<!-- Where $\epsilon$ is a normally distributed random variable with mean $0$ and variance $1$, and $\alpha$ and $\beta$ are constant with respect to the instantaneous time increment $\Delta t$, such that -->

<!-- \begin{equation} -->
<!-- \frac{\Delta v}{v} ~ \phi(\alpha \Delta t, \beta^2 \Delta t) -->
<!-- \end{equation} -->

<!-- Where $\phi()$ is the standard normal probability density function; and hence -->

<!-- \begin{equation} -->
<!-- v(T) = v(\hat{t}) e^{(\alpha - \beta^2 / 2) \tau + \beta \epsilon \sqrt{\tau}} -->
<!-- \label{eq:defVt} -->
<!-- \end{equation} -->

<!-- Where $T$ is some future time step, $\hat{t} < T$ is the time step at which $v(T)$ is evaluated, and $\tau = T - \hat{t}$. (See Hull [@hull9thEdition] for details.) Then, for a constant C, -->
<!-- <!-- , such that --> 
<!-- <!-- \begin{equation} --> 
<!-- <!-- \ln \left(\frac{v(T)}{v(\hat{t})} \right) ~ \phi \left(\left(\alpha - \frac{\beta^2}{2} \right) \tau, \beta^2 \tau \right) -->
<!-- <!-- \end{equation} -->

<!-- \begin{equation} -->
<!-- E[max(v(T) - C, 0)]\bigr|_{t = \hat{t}} = e^{\alpha \tau} v(\hat{t}) \Phi(\delta + \beta \sqrt{\tau}) - C \Phi(\delta) -->
<!-- \end{equation} -->

<!-- Where $\Phi()$ is the standard normal cumulative distribution function, and -->

<!-- \begin{equation} -->
<!-- \delta = \frac{\ln(v(\hat{t}) / C) + (\alpha - \beta^2 / 2) \tau}{\beta \sqrt{\tau}} -->
<!-- \end{equation} -->

<!-- Proof: -->

(The proof below closely follows that of Hull in the appendix to chapter 13 of his book [-@hull9thEdition].)
<!-- _Part 1_ -->

By definition, for a random variable $q$ and a constant $C$,

\begin{equation}
E[\max(q - C, 0)] = \int_{C}^{\infty} (q - C) p(q) \: dq
\label{eq:def}
\end{equation}

Where $p(q)$ is the probability density function of the random variable $q$. If $\ln(q)$ is normally distributed with mean $\nu$ and variance $\omega^2$, then

\begin{equation}
E[q] = e^{\nu + \frac{\omega^2}{2}}
\label{eq:muXT}
\end{equation}

and

\begin{equation}
p(q) = \frac{1}{q \omega} \phi \left(\frac{\ln(q) - \nu}{\omega} \right)
\end{equation}

Where $\phi()$ is the standard normal probability density function.

Now, introducing a change of variables,

\begin{equation}
u = \frac{\ln(q) - \nu}{\omega}
\label{eq:subThis1}
\end{equation}

Such that

\begin{equation}
\frac{du}{d q} = \frac{1}{q \omega}\: \rightarrow \: d q = q \omega du
\label{eq:subThis2}
\end{equation}

The definition in Eq \eqref{eq:def} resolves as follows.

\begin{equation}
\begin{split}
E[\max(q - C, 0)] &= \int_{\frac{\ln(C) - \nu}{\omega}}^{\infty} (e^{u \omega + \nu} - C) \phi(u) \: du \\
&= \int_{\frac{\ln(C) - \nu}{\omega}}^{\infty} e^{u \omega + \nu} \phi(u) \: du - C \int_{\frac{\ln(C) - \nu}{\omega}}^{\infty} \phi(u) \: du \\
&= \frac{1}{\sqrt{2 \pi}} \int_{\frac{\ln(C) - \nu}{\omega}}^{\infty} e^{u \omega + \nu -\frac{u^2}{2}} \: du - C \Phi \left(-\frac{\ln(C) - \nu}{\omega} \right) \\
&= \frac{1}{\sqrt{2 \pi}} \int_{\frac{\ln(C) - \nu}{\omega}}^{\infty} e^{-\frac{(u - \omega)^2}{2} + \nu + \frac{\omega^2}{2}} \: du - C \Phi \left(-\frac{\ln(C) - \nu}{\omega} \right) \\
&= e^{\nu + \frac{\omega^2}{2}} \frac{1}{\sqrt{2 \pi}} \int_{\frac{\ln(C) - \nu}{\omega}}^{\infty} e^{-\frac{(u - \omega)^2}{2}} \: du - C \Phi \left(-\frac{\ln(C) - \nu}{\omega} \right) \\
&= E[q] \Phi \left(- \frac{\ln(C) - \nu}{\omega} + \omega \right) - C \Phi \left(-\frac{\ln(C) - \nu}{\omega} \right)
\end{split}
\end{equation}
<!-- The remaining integral on the right-hand side of the definition resolves as follows. -->
<!-- \begin{equation} -->
<!-- \begin{split} -->
<!-- \int_{\frac{\ln(C) - \nu}{\omega}}^{\infty} e^{u \omega + \nu} \phi(u) \: du &= \frac{1}{\sqrt{2 \pi}} \int_{\frac{\ln(C) - \nu}{\omega}}^{\infty} e^{u \omega + \nu -\frac{u^2}{2}} \: du \\ -->
<!-- &= \frac{1}{\sqrt{2 \pi}} \int_{\frac{\ln(C) - \nu}{\omega}}^{\infty} e^{-1 / 2 (u^2 - 2 u \omega - 2 \nu)} \: du \\ -->
<!-- &= \frac{1}{\sqrt{2 \pi}} \int_{\frac{\ln(C) - \nu}{\omega}}^{\infty} e^{-\frac{(u - \omega)^2}{2} + \nu + \frac{\omega^2}{2}} \: du \\ -->
<!-- &= \frac{1}{\sqrt{2 \pi}} \int_{\frac{\ln(C) - \nu}{\omega}}^{\infty} e^{-\frac{(u - \omega)^2}{2}} e^{\nu + \frac{\omega^2}{2}} \: du \\ -->
<!-- &= e^{\nu + \frac{\omega^2}{2}} \frac{1}{\sqrt{2 \pi}} \int_{\frac{\ln(C) - \nu}{\omega}}^{\infty} e^{-\frac{(u - \omega)^2}{2}} \: du \\ -->
<!-- &= \frac{E[q]}{\sqrt{2 \pi}} \int_{\frac{\ln(C) - \nu}{\omega}}^{\infty} e^{-\frac{(u - \omega)^2}{2}} \: du \\ -->
<!-- &= E[q] \int_{\frac{\ln(C) - \nu}{\omega}}^{\infty} \phi(u - \omega) \: du \\ -->
<!-- &= E[q] \Phi \left(- \frac{\ln(C) - \nu}{\omega} + \omega \right) -->
<!-- \end{split} -->
<!-- \end{equation} -->
<!-- The definition may now be rewritten as follows. -->
<!-- \begin{equation} -->
<!-- E[\max(q - C, 0)] = E[q] \Phi \left(- \frac{\ln(C) - \nu}{\omega} + \omega \right) - C \Phi \left(-\frac{\ln(C) - \nu}{\omega} \right) -->
<!-- \end{equation} -->

Note that Eq \eqref{eq:muXT} can be rearranged into an expression for $\nu$.

\begin{equation}
\nu = \ln(E[q]) - \omega^2 / 2
\end{equation}

Substituting this for $\nu$ in the definition gives

\begin{equation}
E[\max(q - C, 0)] = E[q] \Phi \left(\frac{\ln \left( \frac{E[q]}{C} \right) + \frac{\omega^2}{2}}{\omega} \right) - C \Phi \left(-\frac{\ln \left( \frac{E[q]}{C} \right) - \frac{\omega^2}{2}}{\omega} \right)
\end{equation}

$\blacksquare$

## Derivation of Eq \eqref{eq:rov} from the Black-Scholes PDE {-}

Starting from the Black-Scholes PDE,

\begin{equation}
fr = \frac{\partial f}{\partial x} x r + \frac{\partial f}{\partial t} + \frac{s^2 x^2}{2} \frac{\partial^2 f}{\partial x^2}
\end{equation}

Note that

\begin{equation}
x^2 \frac{\partial^2 f}{\partial x^2} = \frac{\partial^2f}{\partial \ln(x)^2} - \frac{\partial f}{\partial \ln(x)}
\end{equation}

Such that the PDE may be rewritten

\begin{equation}
fr = \frac{\partial f}{\partial \ln(x)} \left(r - \frac{s^2}{2} \right) + \frac{\partial f}{\partial t} + \frac{s^2}{2} \frac{\partial^2f}{\partial \ln(x)^2}
\end{equation}

Defining $f = e^{-r \tau} y$, where $\tau = T - t$ for some future time $T$, and substituting this in for $f$,

\begin{equation}
0 = \frac{\partial y}{\partial \ln(x)} \left(r - \frac{s^2}{2} \right) + \frac{\partial y}{\partial t} + \frac{s^2}{2} \frac{\partial^2 y}{\partial \ln(x)^2}
\end{equation}

Substituting $T - \tau$ for $t$ and expressing the equation in terms of a new function $v(u, \tau)$, where $u = \ln(x) + (r - s^2 / 2) \tau$,

\begin{equation}
0 = \frac{s^2}{2} \frac{\partial^2 v}{\partial u^2} - \frac{\partial v}{\partial t}
\end{equation}

This is the one dimensional heat equation, with fundamental solution

\begin{equation}
v(u, \tau) = \frac{1}{\sqrt{2 \pi s^2 \tau}} \int_{-\infty}^{\infty} v(\ell, 0) e^{-\frac{(u - \ell)^2}{2 s^2 \tau}} \:d \ell
\end{equation}

The boundary condition of interest is $f(x, T) = \max(x(T) - K, 0)$, which in terms of $v(u, \tau)$ is $v(u, 0) = \max(e^u - K, 0)$.

\begin{equation}
\begin{split}
v(u, \tau) &= \frac{1}{s \sqrt{2 \pi \tau}} \int_{\ln(K)}^{\infty} (e^{\ell} - K) e^{-\frac{(u - \ell)^2}{2 s^2 \tau}} \: d \ell \\
&= \frac{1}{s \sqrt{2 \pi \tau}} \int_{\ln(K)}^{\infty} e^{\ell - \frac{(u - \ell)^2}{2 s^2}} \:d \ell - \frac{K}{s \sqrt{2 \pi \tau}} \int_{\ln(K)}^{\infty} e^{-\frac{(u - \ell)^2}{2 s^2 \tau}} \:d \ell
\end{split}
\end{equation}

Introducing the substitution

\begin{equation}
z = \frac{\ell - u}{s\sqrt{\tau}} \rightarrow \: s\sqrt{\tau} dz = d \ell
\end{equation}

The integral simplifies to

\begin{equation}
\begin{split}
v(u, \tau) &= \frac{1}{\sqrt{2 \pi}} \int_{\frac{\ln(K) - u}{s \sqrt{\tau}}}^{\infty} e^{z s \sqrt{\tau} + u - \frac{z^2}{2}} \:dz - \frac{K}{\sqrt{2 \pi}} \int_{\frac{\ln(K) - u}{s \sqrt{\tau}}}^{\infty} e^{-\frac{z^2}{2}} \:d z \\
&= \frac{1}{\sqrt{2 \pi}} \int_{\frac{\ln(K) - u}{s \sqrt{\tau}}}^{\infty} e^{- \frac{\left(z - s \sqrt{\tau} \right)^2}{2} + u + \frac{s^2 \tau}{2}} \:d z - K \Phi \left(- \frac{\ln(K) - u}{s \sqrt{\tau}} \right) \\
&= e^{u + \frac{s^2}{2} \tau} \frac{1}{\sqrt{2 \pi}} \int_{\frac{\ln(K) - u}{s \sqrt{\tau}}}^{\infty}  e^{- \frac{\left(z - s \sqrt{\tau} \right)^2}{2}} \: dz - K \Phi \left(- \frac{\ln(K) - u}{s \sqrt{\tau}} \right) \\
&= e^{u + \frac{s^2}{2} \tau} \Phi \left( \frac{-\ln(K) - u}{s \sqrt{\tau}} + s \sqrt{\tau} \right) - K \Phi \left(- \frac{\ln(K) - u}{s \sqrt{\tau}} \right)
\end{split}
\end{equation}

Expressing everything once more in terms of $y(x, \tau)$, this reduces to

\begin{equation}
y(x, \tau) = e^{r \tau} x(t) \Phi(\delta + s \sqrt{\tau}) - K \Phi(\delta)
\end{equation}

Where

\begin{equation}
\delta = \frac{\ln \left( \frac{x(t)}{K} \right) + \left(r - \frac{s^2}{2} \right) \tau}{s \sqrt{\tau}}
\end{equation}

Multiplying through by the discount factor gives the expression for $f(x, \tau)$.

\begin{equation}
f(x, \tau) = x(t) \Phi(\delta + s \sqrt{\tau}) -  e^{-r \tau} K \Phi(\delta)
\end{equation}

Which is equivalent to Eq \eqref{eq:rov} when $t = 0$, $T = T_0$, $K = K_0$.

$\blacksquare$

